### 2022-04-28 outputs/paper/paper.Rmd
Pushed graphing code for each city with updated abstract and introduction

---
title: "Analysing the effects of COVID-19 on Crime in the United States"
subtitle: "A deep dive into criminal activity during the pandemic"
author: 
  - Nayan Saxena
thanks: "Code and data are available at: https://github.com/the-infiltrator/COVID-19-Crime"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "The COVID-19 pandemic caused the various state-level governments in the United States to issue stay-at-home orders in early 2020. These policy measures resulted in sweeping impacts on the everyday life of people with mostly negative implications.  The more negative implications of the pandemic have generally eclipsed the more positive aspects such as the reduction of crime rates by 37% worldwide [@boman2021global], and also the noticeable drop in carbon emissions [@bauwens2020impact] and elevation of air quality around the world—in turn reducing pollution-related respiratory issues  [@dutheil2020covid]. While crime rates globally have gone down, it is also important to take a granular look at the crime data to observe which types of crimes have been most affected,
and further, examine if there is a significant correlation between crime rates and COVID-19 cases — which is precisely the goal of this study. We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites through time series modeling and also exploratory data analysis focusing on the criminal activity in key US cities such as Chicago,  Los Angeles, Philadelphia, and Seattle.  Each city has been chosen specifically to reflect different American aspects, geographical as well as socio-economic, to enable us to form more generalizable as well as granular conclusions. \\par
 \\textbf{Keywords:} COVID-19 Pandemic, Criminal Behaviour, Time Series Modelling, Correlation, Dataset Creation, Criminal Psychology"

output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib

pdf_document:
  extra_dependencies: ["float"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(plyr)
library(flexdashboard)
library(RSocrata)
library(tidyverse)
library(tsbox) # transform data into time series
library(xts)
library(COVID19) # to get data about covid 19
library(forecast) #ariRI model
library(vars) #VAR and Causality
library(dygraphs)
library(leaflet)
library(htmlwidgets)
#[INCOMPLETE] Graphs in Chicago have been converted here to Plotly HTML Widgets 
# Make some noisily increasing data [Testing Purposes]
set.seed(955)
dat <- data.frame(cond = rep(c("A", "B"), each=10),
                  xvar = 1:20 + rnorm(20,sd=3),
                  yvar = 1:20 + rnorm(20,sd=3))
#### COVID19 DATA ####
#Load Chicago Data
covid19_CH <- covid19("USA", level = 3) |>
  # this cook county contains chicago
  filter(administrative_area_level_3 == "Cook",
         administrative_area_level_2 == "Illinois" ) |>
  # filter out days when confirmed is zero or one
  # becasue when it was 2 for a very long time
  filter(confirmed > 2)

# Load Providence data
covid19_RI <- covid19("USA", level = 2) |>
  filter(administrative_area_level_2 == "Rhode Island") |>
  # filter out days when confirmed is zero or one
  # becasue when it was 1 for a very long time
  filter(confirmed > 1)
## March 07 has 140 confirmed case which is impossible.
## Google shows that date had still 3 cumulative case
## Manual adjustment on row 5
covid19_RI$confirmed[5] = covid19_RI$confirmed[4]

# Load Boston Data
covid19_MA <- covid19("USA", level = 2) |>
  filter(administrative_area_level_2 == "Massachusetts") |>
  # filter out days when confirmed is zero or one
  # becasue when it was 1 for a very long time
  filter(confirmed > 1)

# Load LA data
# extract LA data from US data. 
covid19_LA <- covid19("USA", level = 3) |>
  filter(administrative_area_level_3 == "Los Angeles",
         administrative_area_level_2 == "California") |>
  # stayed at 1 for long time
  filter(confirmed > 1,
         date < "2020-06-12")

# Load Atlanta Data
covid19_AT <- covid19("USA", level = 3) |>
  filter(administrative_area_level_2 == "Georgia",
         administrative_area_level_3 == 'Fulton') |>
  # filter out days when confirmed is zero or one
  # becasue when it was 2 for a very long time
  filter(confirmed > 2)

# Load Seattle Data
covid19_SEA <- covid19("USA", level = 3) |>
  # this cook county contains chicago
  filter(administrative_area_level_3 == "King",
         administrative_area_level_2 == "Washington" ) |>
  # filter out days when confirmed is zero or one
  # becasue when it was 1 for a very long time
  filter(confirmed > 1)

# extract Pennsylvania data from US data. 
covid19_PA <- covid19("USA", level = 3) |>
  filter(administrative_area_level_2 == "Pennsylvania",
         administrative_area_level_3 == "Philadelphia") |>
  # filter out days when confirmed is zero
  filter(confirmed > 0)

chicago<- read.csv2('data/chicago.csv')
la<- read.csv2('data/la.csv')
seattle<- read.csv2('data/seattle.csv')
phil<- read.csv2('data/phil.csv')

# summary of all crimes
chicago_summary <- chicago |>
  group_by(primary_type) |>
  dplyr::summarise(number_of_crime = dplyr::n()) |>
  arrange(desc(number_of_crime))

phil_summary <- phil |>
  group_by(text_general_code) |>
  summarise(number_of_crime = n()) |>
  arrange(desc(number_of_crime))

LA_summary <- la |>
  group_by(crm_cd_desc) |>
  summarise(number_of_crime = n()) |>
  arrange(desc(number_of_crime))

seattle_summary <- seattle |>
  group_by(offense) |>
  dplyr::summarise(number_of_crime = dplyr::n()) |>
  arrange(desc(number_of_crime))
```

# Introduction
The COVID-19 pandemic caused the various state-level governments in the United States to issue stay-at-home orders in early 2020. These policy measures resulted in sweeping impacts on the everyday life of people with mostly negative implications.  The onset of the pandemic brought with it a host of other issues like job losses, unemployment, financial crisis, and mental health issues. The pandemic also forced people to stay at home for prolonged periods, sometimes over a month, which seemed to have exasperated opioid abuse and also relapse of a host of other addictions. Furthermore, it has also been shown that the pandemic has led to increased stress and anxiety levels among individuals [@boman2020has].  These negative implications of the pandemic have generally eclipsed the more positive aspects such as the reduction of crime rates by 37% worldwide [@boman2021global], and also the noticeable drop in carbon emissions [@bauwens2020impact] and elevation of air quality around the world—in turn reducing pollution-related respiratory issues  [@dutheil2020covid]. While crime rates globally have gone down, it is also important to take a granular look at the crime data to observe which types of crimes have been most affected,
and further, examine if there is a significant correlation between crime rates and COVID-19 cases — which is precisely the goal of this study.

The main goal of our paper is to examine the effect that the pandemic has brought in terms of specific crimes such as assault, battery, criminal theft, and much more based on state-level crime data within the United States. The main motivation behind this research is the scarcity of work examining the effect of the pandemic on crime, even though there has been a lot of work done examining other aspects concerning the pandemic. News stations have consistently reported lower crime rates but have based their analyses on simply the rate of 911 calls and not actual police data. We aim to bridge this gap by focusing on the criminal activity in key US cities such as Chicago,  Los Angeles, Philadelphia, and Seattle.  Each city has been chosen specifically to reflect different American aspects, geographical as well as socio-economic, to enable us to form more generalizable as well as granular conclusions.

The key motivation for our work is previous work showing that certain crimes like domestic violence saw a significantly smaller drop as compared to other crimes after the onset of the pandemic [@bullinger2020covid], indicating the importance of the nature of the actual crime committed. Our focus on individual city data is due to known data collection issues for such research [@boman2020has] and is motivated by preliminary studies showing that the crime rates for specific crimes are largely dependent on the city [@ashby2020initial]. There are also other factors such as the George Floyd protests in June 2020 that we hypothesize may have resulted in a noticeable increase in the vandalism and theft crimes in certain cities.

Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global, @campedelli2020disentangling, @nivette2021global] alongside work prompting further research in the area [@eisner2020violence]. We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites through time series modeling and also exploratory data analysis.

In the following section, we outline how data for this study was collected alongside an overview of the data. In Section 3 we outline, in detail our modeling approaches and in Section 4 the results are provided for each city. We conclude our study with a detailed discussion in Section 5, explaining some of our key findings and conclusions.

```{r ,fig.cap = "Summary of several important variables present within the survey data",fig.height = 5, fig.width = 5, out.width = "300px", message = FALSE, echo = FALSE, warning=FALSE, fig.align="center"}
library(mapview)
Location <- c("Los Angeles","Chicago","Seattle","Philadelphia" )
lat <- c(33.82377,41.78798,47.714965, 39.952583)
lng <- c( -118.2668,-87.7738,-122.127166 ,-75.165222)
df <- data.frame(Location, lat,lng)

m <- leaflet() |>
  addTiles() |>  # Add default OpenStreetMap map tiles
  addMarkers(data=df,)
mapshot(m, file = "~/Rplot.png")
knitr::include_graphics("~/Rplot.png")
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
```

# Data
The primary data set used is publicly available from the the R, COVID-19 Data Hub [@cdhub] containing a daily summary of COVID-19 cases, deaths, recovered, tests, vaccinations, and hospitalizations for 230+ countries, 760+ regions, and 12000+ administrative divisions of lower level including policy measures, mobility, and geospatial data. Furthermore, we use city data for each city involved in our study to examine the crime rates across the pandemic and the city-wise data sources are outlined below alongside exploratory analyses of both COVID-19 and crime data.

## Chicago

```{r graph1, fig.cap = "Daily Confirmed Cases of COVID-19 in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = chicago|>mutate(Date=as.Date(Date))

ts_CH <- covid19_CH %>% 
  dplyr::select(date, confirmed) %>%
  ts_xts()

# try first log difference
ts_diff_CH <- na.omit(diff(ts_CH))
plot.xts(ts_diff_CH,
         main = "Daily confirmed cases of COVID-19 in Chicago")

```

```{r graph2, fig.cap = "Yearly Trends for the top five crimes in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
# 2020 only
plt <- copy %>%
  dplyr::select(y_month, month, primary_type, year) %>%
  filter(primary_type %in% chicago_summary$primary_type[1:5] | primary_type == "MOTOR VEHICLE THEFT", y_month != "2020-06") %>%
  dplyr::count(year, month, primary_type) %>%
  na.omit() %>% 
  mutate(Date = as.Date(paste("2000", month, "01", sep = "-"))) %>%
  ggplot(aes(x=Date, y=n, group = year, color = as.character(year))) + 
  geom_line() + 
  facet_wrap(~primary_type, nrow = 1) +
  guides(color = guide_legend(reverse = TRUE)) +
   ylab('') + xlab("")+
  theme(legend.title = element_blank()) +
  scale_x_date(date_labels = "%b")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

# ggplotly(chicago_?SZdaily, tooltip = c("Date", "n"))
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
plt
```

```{r graph3, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
# 2020 only

chicago_daily <- copy |>
  dplyr::select(Date, primary_type, year) |>
  filter(primary_type %in% chicago_summary$primary_type[1:5] | primary_type == "MOTOR VEHICLE THEFT", 
         year == 2020) |>
 dplyr::count(Date, primary_type) |>
  na.omit() |>
  ggplot(aes(Date, n, group = primary_type, color = primary_type)) +
  geom_line() +
  facet_wrap(~primary_type, nrow = 2) +
  scale_fill_brewer(palette = "Set1", breaks = rev(levels(chicago_summary$primary_type[1:5]))) +
  ylab('Number of Reported Crime') + xlab("")+
  theme(legend.position = "none")+
  theme(text = element_text(size=10), 
        plot.title = element_text(hjust = 0.5))
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
chicago_daily
# ggplotly(chicago_?SZdaily, tooltip = c("Date", "n"))

```

## Los Angeles

```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = la|>mutate(date_occ=as.Date(date_occ))

ts_LA <- covid19_LA %>% 
  dplyr::select(date, confirmed) %>%
  ts_xts()

# try first log difference
ts_diff_LA <- na.omit(diff(ts_LA))
plot.xts(ts_diff_LA,
         main = "Daily confirmed cases of COVID-19 in LA")
```

```{r graph6, fig.cap = "Yearly Trends for the top crimes in Los Angeles",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
ty <- copy %>%
  dplyr::select(y_month, month, crm_cd_desc, year) %>%
  filter(crm_cd_desc %in% LA_summary$crm_cd_desc[1:5],
         y_month != "2020-06") %>%
  count(year, month, crm_cd_desc) %>%
  na.omit() %>%
  mutate(Date = as.Date(paste("2000", month, "01", sep = "-"))) %>%
  ggplot(aes(x=Date, y=n,
             group = year,
             color = as.character(year))) +
  geom_line() +
  facet_free(~crm_cd_desc) +
  guides(color = guide_legend(reverse = TRUE)) +
   ylab('') + xlab("")+
  theme(legend.title = element_blank()) +
  scale_x_date(date_labels = "%b")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

# ggplotly(chicago_?SZdaily, tooltip = c("Date", "n"))

ty
```

```{r graph7, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
daily <- copy %>%
  dplyr::select(date_occ, crm_cd_desc, year) %>%
  filter(crm_cd_desc %in% LA_summary$crm_cd_desc[1:5],
         year == 2020) %>%
  count(date_occ, crm_cd_desc) %>%
  ggplot(aes(date_occ, n, group = crm_cd_desc, color = crm_cd_desc)) +
  geom_line() +
  facet_wrap(~crm_cd_desc, nrow=2) +
  scale_fill_brewer(palette = "Set1", breaks = rev(levels(LA_summary$crm_cd_desc[1:5]))) +
  ylab('') + xlab("")+
  theme(legend.position = "none")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

# ggplotly(chicago_?SZdaily, tooltip = c("Date", "n"))

daily
```

## Seattle

```{r graph8, fig.cap = "Daily Confirmed Cases of COVID-19 in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = seattle|>mutate(Date=as.Date(Date))

ts_ST <- covid19_SEA %>% 
  dplyr::select(date, confirmed) %>%
  ts_xts()

# try first log difference
ts_diff_ST <- na.omit(diff(ts_ST))
plot.xts(ts_diff_ST,
         main = "Daily confirmed cases of COVID-19 in Seattle")
```

```{r graph9, fig.cap = "Yearly Trends for the top crimes in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
yty <- copy %>%
  dplyr::select(y_month, MONTH, offense, YEAR) %>%
  filter(offense %in% seattle_summary$offense[1:5],
         y_month != "2020-06") %>%
  dplyr::count(YEAR, MONTH, offense) %>%
  na.omit() %>%
  mutate(Date = as.Date(paste("2000", MONTH, "01", sep = "-"))) %>%
  ggplot(aes(x=Date, y=n, group = YEAR, color = as.character(YEAR))) +
  geom_line() +
  facet_free(~offense, scales = 'free') +
  guides(color = guide_legend(reverse = TRUE)) +
   ylab('') + xlab("")+
  theme(legend.title = element_blank()) +
  scale_x_date(date_labels = "%b") +
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

yty
```

```{r graph10, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")

seattle_daily <- copy %>%
  dplyr::select(Date, offense, YEAR) %>%
  filter(offense %in% seattle_summary$offense[1:5], YEAR=='2020') %>%
  dplyr::count(Date, offense) %>%
  ggplot(aes(Date, n, group = offense, color = offense)) +
  geom_line() +
  facet_wrap(~offense, nrow=2) +
  scale_fill_brewer(palette = "Set1", breaks = rev(levels(seattle_summary$offense[1:5]))) +
  ylab('') + xlab("")+
  theme(legend.position = "none")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))

# ggplotly(chicago_?SZdaily, tooltip = c("Date", "n"))

seattle_daily
```

### Philadelphia

```{r graph11, fig.cap = "Daily Confirmed Cases of COVID-19 in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
copy  = phil|>mutate(date=as.Date(date))

ts_PH <- covid19_PA %>% 
  dplyr::select(date, confirmed) %>%
  ts_xts()

# try first log difference
ts_diff_PH <- na.omit(diff(ts_PH))
plot.xts(ts_diff_PH,
         main = "Daily confirmed cases of COVID-19 in Philadelphia")
```

```{r graph12, fig.cap = "Yearly Trends for the top crimes in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
plt <- phil %>%
  dplyr::select(MONTH, dispatch_date, text_general_code, YEAR, y_month) %>%
  filter(text_general_code %in% phil_summary$text_general_code[1:5], y_month != "2020-06") %>%
  count(YEAR, MONTH, text_general_code) %>%
  na.omit() %>% 
  mutate(Date = as.Date(paste("2000", MONTH, "01", sep = "-"))) %>%
  ggplot(aes(x=Date, y=n, group = YEAR, color = as.character(YEAR))) + 
  geom_line() + 
  facet_wrap(~text_general_code, nrow = 1) +
  guides(color = guide_legend(reverse = TRUE)) +
   ylab('') + xlab("")+
  theme(legend.title = element_blank()) +
  scale_x_date(date_labels = "%b")+
  theme(text = element_text(size=8), 
        plot.title = element_text(hjust = 0.5))
plt
```

```{r graph13, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
# 2020 only
knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
daily <- copy %>%
  dplyr::select(date, text_general_code, YEAR, y_month) %>%
  filter(text_general_code %in% phil_summary$text_general_code[1:5], YEAR == 2020, y_month != '2020-06') %>%
  count(date, text_general_code) %>%
  na.omit() %>%
  ggplot(aes(date, n, group = text_general_code, color = text_general_code))+
  geom_line() + 
  facet_wrap(~text_general_code, nrow=2) +
  scale_fill_brewer(palette = "Set1", breaks = rev(levels(phil_summary$text_general_code[1:5]))) +
  ylab('') + xlab("")+
  theme(legend.position = "none")+
  theme(text = element_text(size=10), 
        plot.title = element_text(hjust = 0.5))
daily

```

\newpage
# Model 
# Results 
# Discussion
## First discussion point 
If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt from all
this.

## Second discussion point 
## Third discussion point 
## Weaknesses and next steps 
Weaknesses and next steps should also be included. 
\newpage 
<!-- \appendix  -->

<!--  # Appendix {-}  -->
<!--  # Additional details  -->

\newpage

# References

### 2022-04-30 outputs/paper/paper.Rmd
Added VAR modelling for each city, impulse response analysis+granger test and also fixed floating of figures.

@@
 bibliography: references.bib
+urlcolor: blue
 
@@
   # stayed at 1 for long time
-  filter(confirmed > 1,
-         date < "2020-06-12")
+  filter(confirmed > 1)
 
@@
 Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global, @campedelli2020disentangling, @nivette2021global] alongside work prompting further research in the area [@eisner2020violence].
-We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites through time series modeling and also exploratory data analysis.
+We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites, using R statistical programming [@citeR], through time series modeling and also exploratory data analysis.
 
@@
 
-```{r ,fig.cap = "Summary of several important variables present within the survey data",fig.height = 5, fig.width = 5, out.width = "300px", message = FALSE, echo = FALSE, warning=FALSE, fig.align="center"}
+```{r ,fig.cap = "Geographical overview of the cities being examined under this study.",fig.height = 5, fig.width = 5, out.width = "300px", message = FALSE, echo = FALSE, warning=FALSE, fig.align="center"}
 library(mapview)
@@
 ## Chicago
+The city of Chicago was chosen for this analysis because it has the highest crime rates in America.
+The gang culture and gun laws have made Chicago the hub of specifically violent crimes, with higher homicide rates on average as compared to other cities.
+Furthermore, during the early stages of the COVID-19 pandemic, it was found that Chicago was one of the first few cities to harbor several virus strains [@anderson], leading to a disaster proclamation by then Governor Pritzker.
+Chicago has since then seen roughly three waves of the virus as shown in Figure \@ref(fig:graph1).
+The data for the city of Chicago was sourced from the [Chicago police](https://home.chicagopolice.org/statistics-data/data-requests/), spanning six main categories of criminal activity which are assault, battery, criminal damage, deceptive practices, vehicle theft, and general theft.
 
-```{r graph1, fig.cap = "Daily Confirmed Cases of COVID-19 in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+```{r graph1, fig.cap = "Daily Confirmed Cases of COVID-19 in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
 # 2020 only
@@
 
+# extract top 5 crime
+top5crime <- chicago %>%
+  filter(primary_type %in% head(chicago_summary$primary_type, 5)) %>%
+  group_by(Date, primary_type) %>%
+  tally() %>%
+  spread(primary_type, n)
+
+# rename columns
+colnames(top5crime) <- c('time',
+                         "assault",
+                         "battery",
+                         "criminal",
+                         'deceptive',
+                         "theft")
+top5crime <- na.omit(top5crime)
+# create time series
+top5crime_xts <- ts_xts(top5crime[,1:2])
+
+for (i in (3:ncol(top5crime))){
+  temp_xts <- ts_xts(top5crime[, c(1,i)])
+  top5crime_xts <- merge(top5crime_xts, temp_xts)
+}
+
+# extract difference, change per day
+top5crime_diff <- na.omit(diff(top5crime_xts))
+
 ```
+When closely examining yearly trends for the different crimes, Figure \@ref(fig:graph2), it can be observed that theft rates and battery noticeably went down in March of 2020 while crimes like assault and Vehicle Theft remained steady.
+Furthermore, it can be observed throughout the pandemic, these levels remained relatively consistent throughout 2021 and 2022.
+In general, it is also noted that most crimes, with the exception of vehicle theft and deception, went up generally in summer and then reduced when the temperatures fell for winter, indicating a seasonal trend.
 
-```{r graph2, fig.cap = "Yearly Trends for the top five crimes in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
+```{r graph2, fig.cap = "Yearly Trends for the top five crimes in Chicago",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
 # 2020 only
@@
 
-```{r graph3, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
-# 2020 only
+Finally, a more granular look at the frequency of each crime only during the pandemic, as shown in Figure \@ref(fig:graph3), shows that most crimes during the pandemic remained relatively consistent with the exception of criminal damage and theft.
+A sudden uptick in criminal damage and theft is noted in June 2020, which can be attributed to the violence caused by certain bad actors during the George Floyd Protests.
+
+```{r graph3, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+# 2020 Onwards only
 
@@
   filter(primary_type %in% chicago_summary$primary_type[1:5] | primary_type == "MOTOR VEHICLE THEFT", 
-         year == 2020) |>
+         year >= 2020) |>
  dplyr::count(Date, primary_type) |>
@@
 
-```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
 # 2020 only
@@
          main = "Daily confirmed cases of COVID-19 in LA")
+
 ```
 
-```{r graph6, fig.cap = "Yearly Trends for the top crimes in Los Angeles",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
-# 2020 only
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
+Proin commodo eros iaculis, faucibus dui et, vehicula felis.
+Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
+Mauris auctor sodales mauris, sed tristique quam sagittis in.
+Mauris ut bibendum enim.
+Maecenas consequat elit ut posuere aliquet.
+In hac habitasse platea dictumst.
+Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+
+```{r graph6, fig.cap = "Yearly Trends for the top crimes in Los Angeles",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
 
-```{r graph7, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r graph7, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
 # 2020 only
@@
   filter(crm_cd_desc %in% LA_summary$crm_cd_desc[1:5],
-         year == 2020) %>%
+         year >= 2020) %>%
   count(date_occ, crm_cd_desc) %>%
@@
 ```
+<!-- \newpage -->
 
@@
 
-```{r graph8, fig.cap = "Daily Confirmed Cases of COVID-19 in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+```{r graph8, fig.cap = "Daily Confirmed Cases of COVID-19 in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
 # 2020 only
@@
 
-```{r graph9, fig.cap = "Yearly Trends for the top crimes in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
-# 2020 only
+```{r graph9, fig.cap = "Yearly Trends for the top crimes in Seattle",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
 
-```{r graph10, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
-# 2020 only
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.
+
+```{r graph10, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
   dplyr::select(Date, offense, YEAR) %>%
-  filter(offense %in% seattle_summary$offense[1:5], YEAR=='2020') %>%
+  filter(offense %in% seattle_summary$offense[1:5], YEAR>='2020') %>%
   dplyr::count(Date, offense) %>%
@@
 
-### Philadelphia
+## Philadelphia
 
-```{r graph11, fig.cap = "Daily Confirmed Cases of COVID-19 in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
-# 2020 only
+```{r graph11, fig.cap = "Daily Confirmed Cases of COVID-19 in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
 
-```{r graph12, fig.cap = "Yearly Trends for the top crimes in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 10,}
-# 2020 only
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.
+Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
+Proin commodo eros iaculis, faucibus dui et, vehicula felis.
+Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
+Mauris auctor sodales mauris, sed tristique quam sagittis in.
+Mauris ut bibendum enim.
+Maecenas consequat elit ut posuere aliquet.
+In hac habitasse platea dictumst.
+Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+
+```{r graph12, fig.cap = "Yearly Trends for the top crimes in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
   dplyr::select(MONTH, dispatch_date, text_general_code, YEAR, y_month) %>%
-  filter(text_general_code %in% phil_summary$text_general_code[1:5], y_month != "2020-06") %>%
+  filter(text_general_code %in% phil_summary$text_general_code[1:5]) %>%
   count(YEAR, MONTH, text_general_code) %>%
@@
 
-```{r graph13, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 6, fig.width = 12,}
-# 2020 only
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r graph13, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+
 knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
@@
   dplyr::select(date, text_general_code, YEAR, y_month) %>%
-  filter(text_general_code %in% phil_summary$text_general_code[1:5], YEAR == 2020, y_month != '2020-06') %>%
+  filter(text_general_code %in% phil_summary$text_general_code[1:5], YEAR >= 2020) %>%
   count(date, text_general_code) %>%
@@
 \newpage
-# Model 
+# Methods
+
+## Vector autoregression Model (VAR)
+
+Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over time.
+VAR is a type of stochastic process model.
+VAR models generalize the single-variable (univariate) autoregressive model by allowing for multivariate time series.
+VAR models are often used in economics and the natural sciences.
+A pth-order VAR is denoted "VAR(p)" and sometimes called "a VAR with p lags". A pth-order VAR model is written as
+
+$${\displaystyle y_{t}=c+A_{1}y_{t-1}+A_{2}y_{t-2}+\cdots +A_{p}y_{t-p}+e_{t},\,}y_{t}=c+A_{1}y_{{t-1}}+A_{2}y_{{t-2}}+\cdots +A_{p}y_{{t-p}}+e_{t},\,$$
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+
+## Granger Causality Test
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.
+Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
+Proin commodo eros iaculis, faucibus dui et, vehicula felis.
+Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
+Mauris auctor sodales mauris, sed tristique quam sagittis in.
+Mauris ut bibendum enim.
+Maecenas consequat elit ut posuere aliquet.
+In hac habitasse platea dictumst.
+Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+
+<!-- \newpage -->
+
+## Forecasting and Impulse Response
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Maecenas tincidunt velit dolor, eu porta lacus ornare et.
+Ut in cursus felis.
+Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
+Aliquam lacinia sagittis lobortis.
+Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
+Morbi aliquet at magna rhoncus vehicula.
+Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
+Sed nisi eros, malesuada nec purus ut, tristique congue dui.
+Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
+Proin commodo eros iaculis, faucibus dui et, vehicula felis.
+Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
+Mauris auctor sodales mauris, sed tristique quam sagittis in.
+Mauris ut bibendum enim.
+Maecenas consequat elit ut posuere aliquet.
+In hac habitasse platea dictumst.
+Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+
+\newpage
 # Results 
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+## Chicago
+
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r top 5 crime VAR, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+
+covid19_CH_diff <- data.frame(diff(covid19_CH$confirmed))
+colnames(covid19_CH_diff)[1] = "confirmed"
+covid19_CH_diff$date = covid19_CH$date[2:length(covid19_CH$date)]
+# time as integer
+covid19_CH_diff$timeInt = as.numeric(covid19_CH_diff$date)
+# make a copy to avoid perfect collinearity
+covid19_CH_diff$timeIid = covid19_CH_diff$timeInt
+
+# make a copy to avoid perfect collinearity
+covid19_CH_diff$timeIid = covid19_CH_diff$timeInt
+
+gamCH <- gamm4::gamm4(confirmed ~  s(timeInt, k=50), random = ~(1|timeIid), 
+	data=covid19_CH_diff, family=poisson(link='log'))
+
+toPredict = data.frame(time = seq(covid19_CH_diff$date[1], 
+                                          covid19_CH_diff$date[length(covid19_CH_diff$date)],
+                                  by = '1 day'))
+toPredict$timeInt = as.numeric(toPredict$time)
+
+# obtain forecast
+forecast <- data.frame(exp(do.call(cbind, mgcv::predict.gam(gamCH$gam, toPredict, se.fit=TRUE))))
+
+# access residuals
+CH_res <- data.frame(covid19_CH_diff$confirmed - forecast$fit)
+
+# transform into time series
+CH_res$time = covid19_CH_diff$date
+colnames(CH_res)[1] = "residuals"
+
+col_order <- c("time", "residuals")
+CH_res <- CH_res[, col_order]
+
+CH_res_ts <- ts_xts(CH_res)
+
+common_time <- seq.Date(start(CH_res_ts), as.Date("2021-12-31"), by = "day")
+
+# combine time series of crime and covid
+combined_diff <- merge(top5crime_diff[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")],
+                       CH_res_ts[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")])
+# plot.xts(CH_res_ts)
+# adf test
+# adf.test(as.ts(CH_res_ts))
+# Stationary process
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+optimal_assault <- VARselect(na.omit(combined_diff)[,c(1,6)], type = 'none', lag.max = 10)
+optimal_battery <- VARselect(na.omit(combined_diff)[,c(2,6)], type = 'none', lag.max = 10)
+optimal_criminal <- VARselect(na.omit(combined_diff)[,c(3,6)], type = 'none', lag.max = 10)
+optimal_deceptive <- VARselect(na.omit(combined_diff)[,c(4,6)], type = 'none', lag.max = 10)
+optimal_theft <- VARselect(na.omit(combined_diff)[,c(5,6)], type = 'none', lag.max = 10)
+
+VAR_assault <- VAR(y=as.ts(na.omit(combined_diff)[,c(1,6)]), p=optimal_assault$selection[1])
+VAR_battery <- VAR(y=as.ts(na.omit(combined_diff)[,c(2,6)]), p=optimal_battery$selection[1])
+VAR_criminal <- VAR(y=as.ts(na.omit(combined_diff)[,c(3,6)]),
+                    p=optimal_criminal$selection[1])
+VAR_deceptive <- VAR(y=as.ts(na.omit(combined_diff)[,c(4,6)]),
+                     p=optimal_deceptive$selection[1])
+VAR_theft <- VAR(y=as.ts(na.omit(combined_diff)[,c(5,6)]), p=optimal_theft$selection[1])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+causality(VAR_assault, cause = colnames(combined_diff)[1])
+causality(VAR_assault, cause = "residuals")
+
+causality(VAR_battery, cause = colnames(combined_diff)[2])
+causality(VAR_battery, cause = "residuals")
+
+causality(VAR_criminal, cause = colnames(combined_diff)[3])
+causality(VAR_criminal, cause = "residuals")
+
+causality(VAR_deceptive, cause = colnames(combined_diff)[4])
+causality(VAR_deceptive, cause = "residuals")
+
+causality(VAR_theft, cause = colnames(combined_diff)[5])
+causality(VAR_theft, cause = "residuals")
+
+vehicle <- chicago %>%
+  filter(primary_type == 'MOTOR VEHICLE THEFT')%>%
+  group_by(date, primary_type) %>%
+  tally() %>%
+  spread(primary_type, n)
+
+colnames(vehicle) <- c('time', 'vehicle')
+vehicle_xts <- ts_xts(na.omit(vehicle)[,1:2])
+vehicle_diff <- na.omit(diff(vehicle_xts))
+
+combined_diff2 <- merge(vehicle_diff[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")],
+                       CH_res_ts[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")])
+optimal_vehicle <- VARselect(na.omit(combined_diff2)[,c(1,2)], type = 'none', lag.max = 10)
+VAR_vehicle <- VAR(y=as.ts(na.omit(combined_diff2)[,c(1,2)]), p=optimal_vehicle$selection[1])
+# 
+# causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
+# causality(VAR_vehicle, cause = "residuals")
+```
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+# use car theft and criminal damage
+par(mfrow = c(1,2))
+lags = c(1:16)
+
+# criminal damange
+# only significant one is from covid to crime
+irf_criminal_2 <- irf(VAR_criminal, 
+                         impulse = "residuals", 
+                         response = "criminal", 
+                         n.ahead = 15,
+                         ortho = F)
+
+# ggplot version of it.
+irf_criminal_2_gg <- data.frame(
+  irf_criminal_2$irf$residuals[,1],
+  irf_criminal_2$Lower$residuals[,1],
+  irf_criminal_2$Upper$residuals[,1]
+)
+
+colnames(irf_criminal_2_gg) <- c("mean", "lower", "upper")
+
+i1 <- ggplot(irf_criminal_2_gg, aes(x = lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new COVID-19 case")+
+  ylab("Criminal Damage")+
+  theme(text = element_text(size=10), 
+        plot.title = element_text(hjust = 0.5))
+i1
+
+interval_value_formatter <- "function(num, opts, seriesName, g, row, col) {
+  value = g.getValue(row, col);
+  if(value[0] != value[2]) {
+    lower = Dygraph.numberValueFormatter(value[0], opts);
+    upper = Dygraph.numberValueFormatter(value[2], opts);
+    return '[' + lower + ', ' + upper + ']';
+  } else {
+    return Dygraph.numberValueFormatter(num, opts);
+  }
+}"
+
+f_criminal <- forecast::forecast(VAR_criminal)
+
+autoplot(f_criminal$forecast$criminal) +
+  ggtitle("Prediction on how many more criminal damage cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+# use car theft and criminal damage
+par(mfrow = c(1,2))
+lags = c(1:16)
+
+# criminal damange
+# only significant one is from covid to crime
+irf_criminal_2 <- irf(VAR_vehicle, 
+                         impulse = "residuals", 
+                         response = "vehicle", 
+                         n.ahead = 15,
+                         ortho = F)
+
+# ggplot version of it.
+irf_criminal_2_gg <- data.frame(
+  irf_criminal_2$irf$residuals[,1],
+  irf_criminal_2$Lower$residuals[,1],
+  irf_criminal_2$Upper$residuals[,1]
+)
+
+colnames(irf_criminal_2_gg) <- c("mean", "lower", "upper")
+
+i1 <- ggplot(irf_criminal_2_gg, aes(x = lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new COVID-19 case")+
+  ylab("Vehicle Theft")+
+  theme(text = element_text(size=10), 
+        plot.title = element_text(hjust = 0.5))
+i1
+
+interval_value_formatter <- "function(num, opts, seriesName, g, row, col) {
+  value = g.getValue(row, col);
+  if(value[0] != value[2]) {
+    lower = Dygraph.numberValueFormatter(value[0], opts);
+    upper = Dygraph.numberValueFormatter(value[2], opts);
+    return '[' + lower + ', ' + upper + ']';
+  } else {
+    return Dygraph.numberValueFormatter(num, opts);
+  }
+}"
+
+f_criminal <- forecast::forecast(VAR_criminal)
+
+autoplot(f_criminal$forecast$criminal) +
+  ggtitle("Prediction on how many more Vehicle Theft cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+## Los Angeles
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r top 5 crime VAR2, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+
+LA<-la
+top5crime <- LA %>%
+  filter(crm_cd_desc %in% head(LA_summary$crm_cd_desc, 5)) %>%
+  group_by(date_occ, crm_cd_desc) %>%
+  tally() %>%
+  spread(crm_cd_desc, n)
+
+# rename columns
+colnames(top5crime) <- c('time',
+                         "battery",
+                         "burglary",
+                         'burglary from vehicle',
+                         "vandalism",
+                         'vehicle')
+# create date
+top5crime$time <- as.Date(top5crime$time,
+                          format = "%Y-%m-%d")
+top5crime <- na.omit(top5crime)
+# create time series
+top5crime_xts <- ts_xts(top5crime[,1:2])
+
+for (i in (3:ncol(top5crime))){
+  temp_xts <- ts_xts(top5crime[, c(1,i)])
+  top5crime_xts <- merge(top5crime_xts, temp_xts)
+}
+
+# extract difference, change per day
+top5crime_diff <- na.omit(diff(top5crime_xts))
+
+ts_LA <- covid19_LA %>% 
+  dplyr::select(date, confirmed) %>%
+  ts_xts()
+
+# try first log difference
+ts_diff_LA <- diff(ts_LA)
+# still clearly not stationary
+# need de-trend
+
+# de-trend 
+# GAMM model from STA303 A3
+
+# calculate the difference per day
+covid19_LA_diff <- data.frame(diff(covid19_LA$confirmed))
+colnames(covid19_LA_diff)[1] = "confirmed"
+covid19_LA_diff$date = covid19_LA$date[2:length(covid19_LA$date)]
+
+# time as integer
+covid19_LA_diff$timeInt = as.numeric(covid19_LA_diff$date)
+# make a copy to avoid perfect collinearity
+covid19_LA_diff$timeIid = covid19_LA_diff$timeInt
+
+# make a copy to avoid perfect collinearity
+covid19_LA_diff$timeIid = covid19_LA_diff$timeInt
+
+# GAMM model
+# 50 too overfit. 15 looks decent
+gamLA <- gamm4::gamm4(confirmed ~  s(timeInt, k=90), random = ~(1|timeIid), 
+	data=covid19_LA_diff, family=poisson(link='log'))
+
+# obtain fitted value
+toPredict = data.frame(time = seq(covid19_LA_diff$date[1], 
+                                          covid19_LA_diff$date[length(covid19_LA_diff$date)],
+                                  by = '1 day'))
+toPredict$timeInt = as.numeric(toPredict$time)
+
+# obtain forecast
+forecast <- data.frame(exp(do.call(cbind, mgcv::predict.gam(gamLA$gam, toPredict, se.fit=TRUE))))
+
+# access residuals
+LA_res <- data.frame(covid19_LA_diff$confirmed - forecast$fit)
+
+# transform into time series
+LA_res$time = covid19_LA_diff$date
+colnames(LA_res)[1] = "residuals"
+
+col_order <- c("time", "residuals")
+LA_res <- LA_res[, col_order]
+
+LA_res_ts <- ts_xts(LA_res)
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+common_time <- seq.Date(start(LA_res_ts), as.Date("2021-12-31"), by = "day")
+
+# combine time series of crime and covid
+combined_diff <- merge(top5crime_diff[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")],
+                       LA_res_ts[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")])
+
+# construct VAR
+optimal_battery <- VARselect(na.omit(combined_diff)[,c(1,6)], type = 'none', lag.max = 10)
+optimal_burglary <- VARselect(na.omit(combined_diff)[,c(2,6)], type = 'none', lag.max = 10)
+optimal_burglary_vehicle <- VARselect(na.omit(combined_diff)[,c(3,6)], type = 'none', lag.max = 10)
+optimal_vandalism <- VARselect(na.omit(combined_diff)[,c(4,6)], type = 'none', lag.max = 10)
+optimal_vehicle <- VARselect(na.omit(combined_diff)[,c(5,6)], type = 'none', lag.max = 10)
+
+VAR_battery <- VAR(y=as.ts(na.omit(combined_diff)[,c(1,6)]),
+                   p=optimal_battery$selection[1])
+VAR_burglary <- VAR(y=as.ts(na.omit(combined_diff)[,c(2,6)]),
+                    p=optimal_burglary$selection[1])
+VAR_burglary_vehicle <- VAR(y=as.ts(na.omit(combined_diff)[,c(3,6)]),
+                    p=optimal_burglary_vehicle$selection[1])
+VAR_vandalism <- VAR(y=as.ts(na.omit(combined_diff)[,c(4,6)]),
+                    p=optimal_vandalism$selection[1])
+VAR_vehicle <- VAR(y=as.ts(na.omit(combined_diff)[,c(5,6)]),
+                   p=optimal_vehicle$selection[1])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+causality(VAR_battery, cause = colnames(combined_diff)[1])
+causality(VAR_battery, cause = "residuals")
+
+causality(VAR_burglary, cause = colnames(combined_diff)[2])
+causality(VAR_burglary, cause = "residuals")
+
+causality(VAR_burglary_vehicle, cause = colnames(combined_diff)[3])
+causality(VAR_burglary_vehicle, cause = "residuals")
+
+causality(VAR_vandalism, cause = colnames(combined_diff)[4])
+causality(VAR_vandalism, cause = "residuals")
+
+causality(VAR_vehicle, cause = colnames(combined_diff)[5])
+causality(VAR_vehicle, cause = "residuals")
+
+# causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
+# causality(VAR_vehicle, cause = "residuals")
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+# use car theft and criminal damage
+lags = c(1:16)
+
+par(mfrow = c(1,2))
+
+# battery
+# criem significant to covid
+irf_battery_1 <- irf(VAR_battery,
+                     impulse = "battery",
+                     response = "residuals",
+                     n.ahead = 15)
+
+# html plot for battery
+irf_battery_1_gg <- data.frame(
+  irf_battery_1$irf$battery[,1],
+  irf_battery_1$Lower$battery[,1],
+  irf_battery_1$Upper$battery[,1]
+)
+
+colnames(irf_battery_1_gg) <- c("mean", "lower", "upper")
+
+irf_battery_1_plot <- ggplot(irf_battery_1_gg, aes(x=lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new Battery case")+
+  ylab("COVID-19")+
+  theme(text = element_text(size=8), 
+        plot.title = element_text(hjust = 0.5))
+
+irf_battery_1_plot
+
+f_battery <- forecast::forecast(VAR_battery)
+
+autoplot(f_battery$forecast$battery) +
+  ggtitle("Prediction on how many more Battery cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+# use car theft and criminal damage
+lags = c(1:16)
+
+par(mfrow = c(1,2))
+
+# battery
+# criem significant to covid
+irf_burglary_1 <- irf(VAR_burglary,
+                     impulse = "burglary",
+                     response = "residuals",
+                     n.ahead = 15)
+
+# html plot for burglary
+irf_burglary_1_gg <- data.frame(
+  irf_burglary_1$irf$burglary[,1],
+  irf_burglary_1$Lower$burglary[,1],
+  irf_burglary_1$Upper$burglary[,1]
+)
+
+colnames(irf_burglary_1_gg) <- c("mean", "lower", "upper")
+
+irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new burglary case")+
+  ylab("COVID-19")+
+  theme(text = element_text(size=8), 
+        plot.title = element_text(hjust = 0.5))
+
+irf_burglary_1_plot
+
+f_burglary <- forecast::forecast(VAR_burglary)
+
+autoplot(f_burglary$forecast$burglary) +
+  ggtitle("Prediction on how many more burglary cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+\newpage
+
+## Philadelphia
+
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r top 5 crime VAR3, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+# extract top 5 crime
+top5crime <- phil %>%
+  filter(text_general_code %in% head(phil_summary$text_general_code, 5)) %>%
+  group_by(dispatch_date, text_general_code) %>%
+  tally() %>%
+  spread(text_general_code, n)
+
+# rename columns
+colnames(top5crime) <- c('time',
+                         "offense",
+                         "assault",
+                         "vehicle",
+                         "thefts",
+                         "vandalism")
+
+# create time series
+top5crime_xts <- ts_xts(top5crime[,1:2])
+
+for (i in (3:ncol(top5crime))){
+  temp_xts <- ts_xts(top5crime[, c(1,i)])
+  top5crime_xts <- merge(top5crime_xts, temp_xts)
+}
+
+# extract difference, change per day
+top5crime_diff <- na.omit(diff(top5crime_xts))
+
+# use loaded covid data 
+# calculate the difference per day
+covid19_PA_diff <- data.frame(diff(covid19_PA$confirmed))
+colnames(covid19_PA_diff)[1] = "confirmed"
+covid19_PA_diff$date = covid19_PA$date[2:length(covid19_PA$date)]
+# extract for tranforming into time series data
+ts_PA <- covid19_PA %>% 
+  dplyr::select(date, confirmed) %>%
+  ts_xts()
+
+# plot time series of PA infection
+#ts_plot(ts_PA)
+# conduct ADF Test
+#adf.test(as.ts(ts_PA))
+# not stationary!
+
+# try first log difference
+ts_diff_PA <- diff(ts_PA)
+#ts_plot(ts_diff_PA)
+# still clearly not stationary
+# need de-trend
+
+# de-trend 
+# GAMM model from STA303 A3
+
+# time as integer
+covid19_PA_diff$timeInt = as.numeric(covid19_PA_diff$date)
+# make a copy to avoid perfect collinearity
+covid19_PA_diff$timeIid = covid19_PA_diff$timeInt
+
+# make a copy to avoid perfect collinearity
+covid19_PA_diff$timeIid = covid19_PA_diff$timeInt
+
+# GAMM model
+# 50 too overfit. 15 looks decent
+#changed to 90
+
+# covid19_PA_diff <- covid19_PA_diff %>%
+#             filter(confirmed > 0)
+covid19_PA_diff$confirmed[covid19_PA_diff$confirmed < 0] <- 0
+
+gamPA <- gamm4::gamm4((confirmed) ~  s(timeInt, k=90), random = ~(1|timeIid), 
+	data=covid19_PA_diff, family=poisson(link='log'))
+
+#fitted value
+toPredict = data.frame(time = seq(covid19_PA_diff$date[1], 
+                                          covid19_PA_diff$date[length(covid19_PA_diff$date)],
+                                  by = '1 day'))
+toPredict$timeInt = as.numeric(toPredict$time)
+
+# obtain forecast
+forecast <- data.frame(exp(do.call(cbind, mgcv::predict.gam(gamPA$gam, toPredict, se.fit=TRUE))))
+
+# access residuals
+PA_res <- data.frame(covid19_PA_diff$confirmed - forecast$fit)
+
+# transform into time series
+PA_res$time = covid19_PA_diff$date
+colnames(PA_res)[1] = "residuals"
+
+col_order <- c("time", "residuals")
+PA_res <- PA_res[, col_order]
+
+PA_res_ts <- ts_xts(PA_res)
+
+# specify common time range
+# start from when covid was a thing
+# end with crime since it is manually updated
+common_time <- seq.Date(start(PA_res_ts), as.Date("2021-12-31"), by = "day")
+
+# combine time series of crime and covid
+combined_diff <- merge(top5crime_diff[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")],
+                       PA_res_ts[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+optimal_offense <- VARselect(na.omit(combined_diff)[,c(1,6)], type = 'none', lag.max = 10)
+optimal_assault <- VARselect(na.omit(combined_diff)[,c(2,6)], type = 'none', lag.max = 10)
+optimal_vehicle <- VARselect(na.omit(combined_diff)[,c(3,6)], type = 'none', lag.max = 10)
+optimal_thefts <- VARselect(na.omit(combined_diff)[,c(4,6)], type = 'none', lag.max = 10)
+optimal_vandalism <- VARselect(na.omit(combined_diff)[,c(5,6)], type = 'none', lag.max = 10)
+
+VAR_offense <- VAR(y=as.ts(na.omit(combined_diff)[,c(1,6)]), p=optimal_offense$selection[1])
+VAR_assault <- VAR(y=as.ts(na.omit(combined_diff)[,c(2,6)]), p=optimal_assault$selection[1])
+VAR_vehicle <- VAR(y=as.ts(na.omit(combined_diff)[,c(3,6)]), p=optimal_vehicle$selection[1])
+VAR_thefts <- VAR(y=as.ts(na.omit(combined_diff)[,c(4,6)]), p=optimal_thefts$selection[1])
+VAR_vandalism <- VAR(y=as.ts(na.omit(combined_diff)[,c(5,6)]), p=optimal_vandalism$selection[1])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+causality(VAR_offense, cause = colnames(combined_diff)[1])
+causality(VAR_offense, cause = "residuals")
+
+causality(VAR_assault, cause = colnames(combined_diff)[2])
+causality(VAR_assault, cause = "residuals")
+
+causality(VAR_vehicle, cause = colnames(combined_diff)[3])
+causality(VAR_vehicle, cause = "residuals")
+
+causality(VAR_thefts, cause = colnames(combined_diff)[4])
+causality(VAR_thefts, cause = "residuals")
+
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+
+lags = c(1:16)
+par(mfrow = c(1,2))
+
+# general offense
+# irf
+irf_offense_1 <- irf(VAR_offense,
+                     impulse = "offense",
+                     response = "residuals",
+                     n.ahead = 15)
+# ggplot
+irf_offense_1_gg <- data.frame(
+  irf_offense_1$irf$offense[,1],
+  irf_offense_1$Lower$offense[,1],
+  irf_offense_1$Upper$offense[,1]
+  )
+
+colnames(irf_offense_1_gg) <- c("mean", "lower", "upper")
+
+irf_offense_1_plot <- ggplot(irf_offense_1_gg, aes(x=lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a New Other Offense Case")+
+  ylab("COVID-19")+
+  theme(text = element_text(size=8), 
+        plot.title = element_text(hjust = 0.5))
+
+irf_offense_1_plot
+
+f_offense <- forecast::forecast(VAR_offense)
+
+autoplot(f_offense$forecast$offense) +
+  ggtitle("Prediction on how many more other offense cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+
+# theft
+# irf
+irf_theft_1 <- irf(VAR_thefts,
+                   impulse = "thefts",
+                   response = "residuals",
+                   n.ahead = 15)
+
+# ggplot
+irf_theft_1_gg <- data.frame(
+  irf_theft_1$irf$thefts[,1],
+  irf_theft_1$Lower$thefts[,1],
+  irf_theft_1$Upper$thefts[,1]
+)
+
+colnames(irf_theft_1_gg) <- c("mean", "lower", "upper")
+
+irf_theft_1_plot <- ggplot(irf_theft_1_gg, aes(x=lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new Theft case")+
+  ylab("COVID-19")+
+  theme(text = element_text(size=8), 
+        plot.title = element_text(hjust = 0.5))
+
+irf_theft_1_plot
+
+f_theft <- forecast::forecast(VAR_thefts)
+
+autoplot(f_theft$forecast$thefts) +
+  ggtitle("Prediction on how many more thefts compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+## Seattle 
+Nulla consectetur erat arcu, eu porta urna sagittis nec.
+In sed vestibulum turpis.
+Nullam vel libero eu sapien eleifend condimentum.
+Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
+Quisque sit amet porttitor mauris, faucibus maximus arcu.
+Duis et lorem eu elit consectetur posuere.
+Aliquam blandit nisl nisl.
+Proin velit arcu, gravida vel tellus.
+
+```{r top 5 crime VAR4, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+# extract top 5 crime
+top5crime <- seattle %>%
+  filter(offense %in% head(seattle_summary$offense, 5)) %>%
+  group_by(Date, offense) %>%
+  tally() %>%
+  spread(offense, n)
+
+# rename columns
+colnames(top5crime) <- c('time',
+                         "larceny",
+                         "burglary",
+                         "vandalism",
+                         'vehicle_theft',
+                         "theft_from_vehicle")
+top5crime <- na.omit(top5crime)
+# create time series
+top5crime_xts <- ts_xts(top5crime[,1:2])
+
+for (i in (3:ncol(top5crime))){
+  temp_xts <- ts_xts(top5crime[, c(1,i)])
+  top5crime_xts <- merge(top5crime_xts, temp_xts)
+}
+
+# extract difference, change per day
+top5crime_diff <- na.omit(diff(top5crime_xts))
+# plot cumulative cases
+# extract for transforming into time series
+ts_SEA <- covid19_SEA %>%
+  dplyr::select(date, confirmed) %>%
+  ts_xts()
+
+# plot daily cases
+# first difference
+ts_diff_SEA <- na.omit(diff(ts_SEA))
+
+covid19_SEA_diff <- data.frame(diff(covid19_SEA$confirmed) - min(ts_diff_SEA))
+
+colnames(covid19_SEA_diff)[1] = "confirmed"
+covid19_SEA_diff$date = covid19_SEA$date[2:length(covid19_SEA$date)]
+
+# time as integer
+covid19_SEA_diff$timeInt = as.numeric(covid19_SEA_diff$date)
+# RIke a copy to avoid perfect collinearity for mixed effect
+covid19_SEA_diff$timeIid = covid19_SEA_diff$timeInt
+
+# GAMM model
+gamSEA <- gamm4::gamm4(confirmed ~ s(timeInt, k = 90), 
+                      random = ~(1|timeIid),
+                      data = covid19_SEA_diff,
+                      family = poisson(link = 'log'))
+
+# plot fitted value
+toPredict = data.frame(time = seq(covid19_SEA_diff$date[1],
+                                  covid19_SEA_diff$date[length(covid19_SEA_diff$date)],
+                                  by = '1 day'))
+
+toPredict$timeInt = as.numeric(toPredict$time)
+forecast_covid <- data.frame(exp(do.call(cbind, mgcv::predict.gam(gamSEA$gam, toPredict, se.fit=TRUE))))
+
+# access residuals
+SEA_res <- data.frame(covid19_SEA_diff$confirmed - forecast_covid$fit)
+
+# transform into time series
+SEA_res$time = covid19_SEA_diff$date
+colnames(SEA_res)[1] = "residuals"
+
+col_order <- c("time", "residuals")
+SEA_res <- SEA_res[, col_order]
+
+SEA_res_ts <- ts_xts(SEA_res)
+
+# start from when covid was a thing
+# end with 1 day before today's date
+common_time <- seq.Date(start(SEA_res_ts), as.Date("2021-12-31") , by = "day")
+
+# combine time series of crime and covid
+combined_diff <- merge(top5crime_diff[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")],
+                       SEA_res_ts[paste(common_time[1],
+                                            common_time[length(common_time)],
+                                            sep = "/")])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+# variable selection based on AIC
+optimal_larceny <- VARselect(na.omit(combined_diff)[,c(1,6)], type = 'none', lag.max = 10)
+optimal_burglary <- VARselect(na.omit(combined_diff)[,c(2,6)], type = 'none', lag.max = 10)
+optimal_vandalism <- VARselect(na.omit(combined_diff)[,c(3,6)], type = 'none', lag.max = 10)
+optimal_vehicle_theft <- VARselect(na.omit(combined_diff)[,c(4,6)], type = 'none', lag.max = 10)
+optimal_theft_fromvehicle <- VARselect(na.omit(combined_diff)[,c(5,6)], type = 'none', lag.max = 10)
+
+# use AIC as selection criteria
+VAR_larceny <- VAR(y=as.ts(na.omit(combined_diff)[,c(1,6)]),
+                   p=optimal_larceny$selection[1])
+VAR_burglary <- VAR(y=as.ts(na.omit(combined_diff)[,c(2,6)]),
+                     p=optimal_burglary$selection[1])
+VAR_vandalism <- VAR(y=as.ts(na.omit(combined_diff)[,c(3,6)]),
+                    p=optimal_vandalism$selection[1])
+VAR_vehicle_theft <- VAR(y=as.ts(na.omit(combined_diff)[,c(4,6)]),
+                         p=optimal_vehicle_theft$selection[1])
+VAR_theft_fromvehicle<- VAR(y=as.ts(na.omit(combined_diff)[,c(5,6)]),
+                              p=optimal_theft_fromvehicle$selection[1])
+```
+
+```{r, message = FALSE, echo = FALSE, warning=FALSE, include=FALSE}
+causality(VAR_larceny, cause = colnames(combined_diff)[1])
+causality(VAR_larceny, cause = "residuals")
+causality(VAR_burglary, cause = colnames(combined_diff)[2])
+causality(VAR_burglary, cause = "residuals")
+causality(VAR_vandalism, cause = colnames(combined_diff)[3])
+causality(VAR_vandalism, cause = "residuals")
+causality(VAR_vehicle_theft, cause = colnames(combined_diff)[4])
+causality(VAR_vehicle_theft, cause = "residuals")
+causality(VAR_theft_fromvehicle, cause = colnames(combined_diff)[5])
+causality(VAR_theft_fromvehicle, cause = "residuals")
+
+```
+
+```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
+Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+lags = c(1:16)
+
+par(mfrow = c(1,2))
+# only covid significant to bulglary
+irf_burglary_1 <- irf(VAR_burglary,
+                      impulse = "residuals",
+                      response = "burglary",
+                      n.ahead = 15)
+# ggplot
+irf_burglary_1_gg <- data.frame(irf_burglary_1$irf$residuals[,1],
+                                irf_burglary_1$Lower$residuals[,1],
+                                irf_burglary_1$Upper$residuals[,1])
+
+colnames(irf_burglary_1_gg) <- c("mean", "lower", "upper")
+
+irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
+  theme_classic() +
+  ggtitle("How many more burglary cases per day there will be 
+          after 1 confirmed covid19 case") +
+  xlab("Number of days after a confimed COVID-19 case")+
+  ylab("Number of bulglary cases")
+
+f_offense <- forecast::forecast(VAR_offense)
+
+irf_burglary_1_plot
+
+f_burglary <- forecast::forecast(VAR_burglary)
+
+autoplot(f_burglary$forecast$burglary) +
+  ggtitle("Prediction on how many more burglarys compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+\newpage
 # Discussion
-## First discussion point 
-If my paper were 10 pages, then should be be at least 2.5 pages. The
-discussion is a chance to show off what you know and what you learnt from all
-this.
 
-## Second discussion point 
-## Third discussion point 
-## Weaknesses and next steps 
-Weaknesses and next steps should also be included. 
+Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris ultricies felis in neque hendrerit semper.
+Sed quis ullamcorper nulla.
+Donec a mauris ut ipsum interdum tempor a et urna.
+Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
+Ut ullamcorper erat lorem, id congue ex ultricies ac.
+Nunc fermentum neque orci.
+Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
+Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
+Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
+Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
+Cras quam mauris, mollis a cursus non, cursus id neque.
+Cras eget elit posuere, ornare dui non, posuere neque.
+
+Ut tempor vestibulum augue sit amet hendrerit.
+Morbi rhoncus dignissim velit.
+Aliquam erat volutpat.
+Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
+Donec sit amet nibh in risus dictum venenatis.
+Mauris non volutpat risus, vitae ornare dui.
+Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
+Donec nec luctus quam.
+Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
+Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
+Duis suscipit sem at enim commodo eleifend.
+Aenean eu malesuada justo.
+Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
+
+Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
+Nam commodo lectus odio, nec hendrerit purus consequat et.
+Sed vel iaculis ex.
+Maecenas mollis nisi id est pharetra facilisis.
+Fusce et tempor lorem.
+Etiam at facilisis nulla.
+Praesent sed erat odio.
+Integer sed dictum tellus, in finibus odio.
+Duis et rhoncus justo.
+Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris ultricies felis in neque hendrerit semper.
+Sed quis ullamcorper nulla.
+Donec a mauris ut ipsum interdum tempor a et urna.
+Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
+Ut ullamcorper erat lorem, id congue ex ultricies ac.
+Nunc fermentum neque orci.
+Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
+Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
+Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
+Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
+Cras quam mauris, mollis a cursus non, cursus id neque.
+Cras eget elit posuere, ornare dui non, posuere neque.
+
+Ut tempor vestibulum augue sit amet hendrerit.
+Morbi rhoncus dignissim velit.
+Aliquam erat volutpat.
+Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
+Donec sit amet nibh in risus dictum venenatis.
+Mauris non volutpat risus, vitae ornare dui.
+Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
+Donec nec luctus quam.
+Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
+Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
+Duis suscipit sem at enim commodo eleifend.
+Aenean eu malesuada justo.
+Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
+
+Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
+Nam commodo lectus odio, nec hendrerit purus consequat et.
+Sed vel iaculis ex.
+Maecenas mollis nisi id est pharetra facilisis.
+Fusce et tempor lorem.
+Etiam at facilisis nulla.
+Praesent sed erat odio.
+Integer sed dictum tellus, in finibus odio.
+Duis et rhoncus justo.
+Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Mauris ultricies felis in neque hendrerit semper.
+Sed quis ullamcorper nulla.
+Donec a mauris ut ipsum interdum tempor a et urna.
+Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
+Ut ullamcorper erat lorem, id congue ex ultricies ac.
+Nunc fermentum neque orci.
+Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
+Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
+Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
+Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
+Cras quam mauris, mollis a cursus non, cursus id neque.
+Cras eget elit posuere, ornare dui non, posuere neque.
+
+Ut tempor vestibulum augue sit amet hendrerit.
+Morbi rhoncus dignissim velit.
+Aliquam erat volutpat.
+Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
+Donec sit amet nibh in risus dictum venenatis.
+Mauris non volutpat risus, vitae ornare dui.
+Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
+Donec nec luctus quam.
+Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
+Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
+Duis suscipit sem at enim commodo eleifend.
+Aenean eu malesuada justo.
+Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
+
+Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
+Nam commodo lectus odio, nec hendrerit purus consequat et.
+Sed vel iaculis ex.
+Maecenas mollis nisi id est pharetra facilisis.
+Fusce et tempor lorem.
+Etiam at facilisis nulla.
+Praesent sed erat odio.
+Integer sed dictum tellus, in finibus odio.
+Duis et rhoncus justo.Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+
+Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
+Nam commodo lectus odio, nec hendrerit purus consequat et.
+Sed vel iaculis ex.
+Maecenas mollis nisi id est pharetra facilisis.
+Fusce et tempor lorem.
+Etiam at facilisis nulla.
+Praesent sed erat odio.
+Integer sed dictum tellus, in finibus odio.
+Duis et rhoncus justo.Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+
+Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
+Nam commodo lectus odio, nec hendrerit purus consequat et.
+Sed vel iaculis ex.
+Maecenas mollis nisi id est pharetra facilisis.
+Fusce et tempor lorem.
+Etiam at facilisis nulla.
+Praesent sed erat odio.
+Integer sed dictum tellus, in finibus odio.
+Duis et rhoncus justo.Pellentesque id purus eros.
+Curabitur rutrum ut nisl vel facilisis.
+Maecenas dapibus vel justo ut aliquet.
+Suspendisse at auctor sem.
+Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
+Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+
 \newpage 

### 2022-04-30 scripts/01-data_cleaning.R

@@ -86,21 +86,20 @@ covid19_PA <- covid19("USA", level = 3) %>%
   filter(confirmed > 0)
 
 
-
-
-
 ########################################################################
 ############################# PHILLY @##################################
 ########################################################################
-
+phil2021 <- read.csv("inputs/data/phil_2021.csv")
+phil2021 <- subset (phil2021, select = -c( the_geom , cartodb_id, the_geom_webmercator ))
 phil2020 <- read.csv("inputs/data/phil_2020.csv")
+phil2020 <- subset (phil2020, select = -c( the_geom , cartodb_id, the_geom_webmercator ))
 phil2019 <- read.csv("inputs/data/phil_2019.csv")
 phil2018 <- read.csv("inputs/data/phil_2018.csv")
 phil2017 <- read.csv("inputs/data/phil_2017.csv")
 phil2016 <- read.csv("inputs/data/phil_2016.csv")
 phil2015 <- read.csv("inputs/data/phil_2015.csv")
 
-phil <- do.call("rbind", list(phil2020, phil2019, phil2018, phil2017, phil2016, phil2015))
+phil <- do.call("rbind", list(phil2021,phil2020, phil2019, phil2018, phil2017, phil2016, phil2015))
 
 # add YEAR, MONTH, y_month
 phil <- phil %>%

### 2022-04-30 outputs/paper/paper.Rmd
Impulse response, granger causality and fixed floating.

@@
 Furthermore, it can be observed throughout the pandemic, these levels remained relatively consistent throughout 2021 and 2022.
-In general, it is also noted that most crimes, with the exception of vehicle theft and deception, went up generally in summer and then reduced when the temperatures fell for winter, indicating a seasonal trend.
 
@@
 ```
-
+In general, it is also noted that most crimes, with the exception of vehicle theft and deception, went up generally in summer and then reduced when the temperatures fell for winter, indicating a seasonal trend.
 Finally, a more granular look at the frequency of each crime only during the pandemic, as shown in Figure \@ref(fig:graph3), shows that most crimes during the pandemic remained relatively consistent with the exception of criminal damage and theft.
@@
 
-```{r graph3, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+```{r graph3, fig.cap = "Frequency of each crime in Chicago during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
 # 2020 Onwards only
@@
 ```
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
-Proin commodo eros iaculis, faucibus dui et, vehicula felis.
-Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
-Mauris auctor sodales mauris, sed tristique quam sagittis in.
-Mauris ut bibendum enim.
-Maecenas consequat elit ut posuere aliquet.
-In hac habitasse platea dictumst.
-Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+Los Angeles, located in California, has historically been a hub of criminal activities mainly centered around narcotics due to its proximity to Tijuana and in general the US-Mexico border.
+Crime in Los Angeles is also prevalent due to the presence of one of the largest populations of homeless people in the United States mostly within the Skid-Row locality.
+Los Angeles was at the peak of criminal activity between 1970 and 1990, and more recently in 2015, it was revealed that the Los Angeles Police were under-reporting criminal activity for roughly 8 years, artificially lowering the crime rates and making it an intriguing city to examine from a data perspective [@bird2018impact].
+To examine the crime rates within the city we source data directly from the  [Los Angeles Police Department Socrata API](https://dev.socrata.com/foundry/data.lacity.org/2nrs-mtv8) and look at the top five crimes which are mainly variants of burglary and theft crimes, alongside vehicle theft and assault.
+Finally, as shown in Figure \@ref(fig:graph5), the pandemic had a significant impact in Los Angeles with some of the first patients of COVID-19, including "patient zero", being from Los Angeles.
+California overall also faced a severe health crisis with cases up to ten times more than were anticipated.
+
+Further examining yearly trends for several crimes, based on Figures \@ref(fig:graph6) and \@ref(fig:graph7), it was found that several crimes like battery/assault, vehicle theft, and burglary increased.
+The noticeable increase in these crimes remained consistent for years spanning the pandemic, although a noticeable drop in battery, burglary, and vehicle theft was observed in January 2022.
 
@@
 
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
-
-```{r graph7, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+```{r graph7, fig.cap = "Frequency of each crime in LA during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
 # 2020 only

### 2022-04-30 outputs/paper/paper.Rmd
Added text for Seattle and Philly with details on data sources and overview of exploratory analysis.

@@
 
-```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 4, fig.width = 8,}
+```{r graph5, fig.cap = "Daily Confirmed Cases of COVID-19 in L.A",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 3, fig.width = 8,}
 # 2020 only
@@
 ```
+\newpage
+
 Los Angeles, located in California, has historically been a hub of criminal activities mainly centered around narcotics due to its proximity to Tijuana and in general the US-Mexico border.
@@
 Los Angeles was at the peak of criminal activity between 1970 and 1990, and more recently in 2015, it was revealed that the Los Angeles Police were under-reporting criminal activity for roughly 8 years, artificially lowering the crime rates and making it an intriguing city to examine from a data perspective [@bird2018impact].
+
 To examine the crime rates within the city we source data directly from the  [Los Angeles Police Department Socrata API](https://dev.socrata.com/foundry/data.lacity.org/2nrs-mtv8) and look at the top five crimes which are mainly variants of burglary and theft crimes, alongside vehicle theft and assault.
-Finally, as shown in Figure \@ref(fig:graph5), the pandemic had a significant impact in Los Angeles with some of the first patients of COVID-19, including "patient zero", being from Los Angeles.
+Finally, as shown in Figure \@ref(fig:graph5), we note that the pandemic had a significant impact in Los Angeles with some of the first patients of COVID-19, including "patient zero", being from Los Angeles.
 California overall also faced a severe health crisis with cases up to ten times more than were anticipated.
@@
 ```
-<!-- \newpage -->
+\newpage
 
@@
 
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.
-
-```{r graph10, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+One of the fastest-growing cities in the United States, Seattle, was specifically chosen for this study due to its high rates of property crime while having relatively constant overall crime rates across the decades.
+As shown in Figure \@ref(fig:graph8), the city was relatively less impacted by COVID-19, making it an interesting candidate in contrast to other cities when it comes to observing criminal behavior.
+Seattle is also the location of the earliest clinical trial being conducted for the Moderna COVID-19 vaccine which occurred around March 20, 2020 [@tuesday_2020], and more recently has seen a sudden increase in criminal activity alongside increased homelessness [@westneat_2022].
+
+ The crime data for Seattle was sourced from the public [Seattle Police Department Records](https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5), with our focus mainly being on the top crimes comprising assault, burglary, vandalism, larceny and theft from a vehicle.
+Given the relatively low rate of COVID-19 in Seattle, trends across most crimes seemed to remain constant before and after the onset of the pandemic as seen in Figure \@ref(fig:graph9).
+It was also observed that crimes such as burglary increased while theft from motor vehicles showed a decline as compared to previous trends.
+These relatively consistent trends during the pandemic are also visible in Figure \@ref(fig:graph10), with an increase in larceny, burglary, and vandalism crimes noted in June-July 2020 due to the Goerge Floyd protests.
+
+```{r graph10, fig.cap = "Frequency of each crime in Seattle during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
 
@@
 ```
+The last city in our study is Philadelphia, which is one of the most violent cities in America along with the highest poverty rates.
+The city is the sixth most populous city in the United States with a rich cultural diversity making it ideal for our study.
+The city was significantly affected by the pandemic as noted by its poor economic performance and decline in the supply of workers [@haider_eichel].
+It should be noted though that when compared to the other cities in our study, Philadelphia has the lowest number of daily cases on average as shown in Figure Figure \@ref(fig:graph11).
+
+To examine the impact on crime due to the pandemic we sourced data from the [Philadelphia Police Department](https://www.phillypolice.com/crime-maps-stats/), with an emphasis on the top crimes which were basically assaults, other offenses, thefts from vehicles, general thefts and vandalism.
+Overall it was found, as demonstrated in Figure \@ref(fig:graph12) that petty offenses increased during the pandemic while most other crimes remained relatively consistent with patterns emerging in prior years.
+An immediate downtrend was also observed in crimes such as assaults and thefts right after the onset of the pandemic.
+
+```{r graph12, fig.cap = "Yearly Trends for the top crimes in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
+
+knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
+plt <- phil %>%
+  dplyr::select(MONTH, dispatch_date, text_general_code, YEAR, y_month) %>%
+  filter(text_general_code %in% phil_summary$text_general_code[1:5]) %>%
+  count(YEAR, MONTH, text_general_code) %>%
+  na.omit() %>% 
+  mutate(Date = as.Date(paste("2000", MONTH, "01", sep = "-"))) %>%
+  ggplot(aes(x=Date, y=n, group = YEAR, color = as.character(YEAR))) + 
+  geom_line() + 
+  facet_wrap(~text_general_code, nrow = 1) +
+  guides(color = guide_legend(reverse = TRUE)) +
+   ylab('') + xlab("")+
+  theme(legend.title = element_blank()) +
+  scale_x_date(date_labels = "%b")+
+  theme(text = element_text(size=8), 
+        plot.title = element_text(hjust = 0.5))
+plt
+```
+
+Further examining the crime trends during years spanning the pandemic, as shown in Figure \@ref(fig:graph13), we observe that there is a general downtrend in petty offenses while most other crimes remain consistent with some seasonal trends in assaults and thefts.
+Finally, as noted previously for other cities, an uptick in criminal activity is observed during the black lives matter protests of June 2020.
+
+```{r graph13, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
+
+knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
+daily <- copy %>%
+  dplyr::select(date, text_general_code, YEAR, y_month) %>%
+  filter(text_general_code %in% phil_summary$text_general_code[1:5], YEAR >= 2020) %>%
+  count(date, text_general_code) %>%
+  na.omit() %>%
+  ggplot(aes(date, n, group = text_general_code, color = text_general_code))+
+  geom_line() + 
+  facet_wrap(~text_general_code, nrow=2) +
+  scale_fill_brewer(palette = "Set1", breaks = rev(levels(phil_summary$text_general_code[1:5]))) +
+  ylab('') + xlab("")+
+  theme(legend.position = "none")+
+  theme(text = element_text(size=10), 
+        plot.title = element_text(hjust = 0.5))
+daily
+
+```
+
+\newpage
+# Methods
+
+## Vector autoregression Model (VAR)
+
+Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over time.
+VAR is a type of stochastic process model.
+VAR models generalize the single-variable (univariate) autoregressive model by allowing for multivariate time series.
+VAR models are often used in economics and the natural sciences.
+A pth-order VAR is denoted "VAR(p)" and sometimes called "a VAR with p lags". A pth-order VAR model is written as
+
+$${\displaystyle y_{t}=c+A_{1}y_{t-1}+A_{2}y_{t-2}+\cdots +A_{p}y_{t-p}+e_{t},\,}y_{t}=c+A_{1}y_{{t-1}}+A_{2}y_{{t-2}}+\cdots +A_{p}y_{{t-p}}+e_{t},\,$$
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec egestas justo in nulla bibendum ultrices.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Donec porttitor interdum purus, vel commodo lorem lacinia at.
+Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
+Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
+Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+
+## Granger Causality Test
 
@@
 
-```{r graph12, fig.cap = "Yearly Trends for the top crimes in Philadelphia",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 2, fig.width = 9,}
-
-knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
-plt <- phil %>%
-  dplyr::select(MONTH, dispatch_date, text_general_code, YEAR, y_month) %>%
-  filter(text_general_code %in% phil_summary$text_general_code[1:5]) %>%
-  count(YEAR, MONTH, text_general_code) %>%
-  na.omit() %>% 
-  mutate(Date = as.Date(paste("2000", MONTH, "01", sep = "-"))) %>%
-  ggplot(aes(x=Date, y=n, group = YEAR, color = as.character(YEAR))) + 
-  geom_line() + 
-  facet_wrap(~text_general_code, nrow = 1) +
-  guides(color = guide_legend(reverse = TRUE)) +
-   ylab('') + xlab("")+
-  theme(legend.title = element_blank()) +
-  scale_x_date(date_labels = "%b")+
-  theme(text = element_text(size=8), 
-        plot.title = element_text(hjust = 0.5))
-plt
-```
-
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
-
-```{r graph13, fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.",message = FALSE, echo = FALSE, warning=FALSE,fig.height = 5, fig.width = 10,}
-
-knitr::opts_chunk$set(fig.pos = "!htbp", out.extra = "")
-daily <- copy %>%
-  dplyr::select(date, text_general_code, YEAR, y_month) %>%
-  filter(text_general_code %in% phil_summary$text_general_code[1:5], YEAR >= 2020) %>%
-  count(date, text_general_code) %>%
-  na.omit() %>%
-  ggplot(aes(date, n, group = text_general_code, color = text_general_code))+
-  geom_line() + 
-  facet_wrap(~text_general_code, nrow=2) +
-  scale_fill_brewer(palette = "Set1", breaks = rev(levels(phil_summary$text_general_code[1:5]))) +
-  ylab('') + xlab("")+
-  theme(legend.position = "none")+
-  theme(text = element_text(size=10), 
-        plot.title = element_text(hjust = 0.5))
-daily
-
-```
-
-\newpage
-# Methods
-
-## Vector autoregression Model (VAR)
-
-Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over time.
-VAR is a type of stochastic process model.
-VAR models generalize the single-variable (univariate) autoregressive model by allowing for multivariate time series.
-VAR models are often used in economics and the natural sciences.
-A pth-order VAR is denoted "VAR(p)" and sometimes called "a VAR with p lags". A pth-order VAR model is written as
-
-$${\displaystyle y_{t}=c+A_{1}y_{t-1}+A_{2}y_{t-2}+\cdots +A_{p}y_{t-p}+e_{t},\,}y_{t}=c+A_{1}y_{{t-1}}+A_{2}y_{{t-2}}+\cdots +A_{p}y_{{t-p}}+e_{t},\,$$
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-
-## Granger Causality Test
+<!-- \newpage -->
+
+## Forecasting and Impulse Response
 
@@
 
-<!-- \newpage -->
-
-## Forecasting and Impulse Response
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.
-Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
-Proin commodo eros iaculis, faucibus dui et, vehicula felis.
-Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
-Mauris auctor sodales mauris, sed tristique quam sagittis in.
-Mauris ut bibendum enim.
-Maecenas consequat elit ut posuere aliquet.
-In hac habitasse platea dictumst.
-Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
-
 \newpage

### 2022-05-01 outputs/paper/paper.Rmd
Added complete results section with analysis of Granger tests

@@
 
+Next, we take a look at the modelling approaches employed to further verify correlations between COVID-19 and crime data.
+
 \newpage
+
 # Methods
@@
 \newpage
+
 # Results 
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
+We now employ the methods outlined in the previous section to narrow down on the crimes that are the most significantly correlated with COVID-19 activity in the respective cities.
+For each city, the impulse response alongside the model predictions are visualized for the crimes that are most significantly associated with COVID-19.
 
@@
 
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
+Based on the fitted VAR models for top crimes in Chicago, we find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as criminal damage ($p=0.8271$), and battery ($p=0.63$).
+Whereas, a statistically significant instantaneous correlation was found between COVID-19 cases impacting assault ($p=0.036$) and a significant Granger-causality with the cases impacting vehicle theft ($p = 0.00022$).
+The impulse response of a new case to these crimes and the model predictions are outlined in Figures \@ref(fig:r11) & \@ref(fig:r12).
 
@@
 # 
-# causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
-# causality(VAR_vehicle, cause = "residuals")
-```
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+causality(VAR_vehicle, cause = colnames(combined_diff2)[1])
+causality(VAR_vehicle, cause = "residuals")
+```
+
+```{r r11, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on assault (left), and VAR model prediction for assault (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
 # use car theft and criminal damage
@@
 # only significant one is from covid to crime
-irf_criminal_2 <- irf(VAR_criminal, 
+irf_assault_2 <- irf(VAR_assault, 
                          impulse = "residuals", 
-                         response = "criminal", 
+                         response = "assault", 
+                         n.ahead = 15,
+                         ortho = F)
+
+# ggplot version of it.
+irf_assault_2_gg <- data.frame(
+  irf_assault_2$irf$residuals[,1],
+  irf_assault_2$Lower$residuals[,1],
+  irf_assault_2$Upper$residuals[,1]
+)
+
+colnames(irf_assault_2_gg) <- c("mean", "lower", "upper")
+
+i1 <- ggplot(irf_assault_2_gg, aes(x = lags)) +
+  geom_line(aes(y = mean), color = "black") +
+  geom_hline(yintercept = 0, color = "blue") +
+  theme_classic() +
+  ggtitle("") +
+  xlab("Days after a new COVID-19 case")+
+  ylab("Assualt")+
+  theme(text = element_text(size=10), 
+        plot.title = element_text(hjust = 0.5))
+i1
+
+f_assault <- forecast::forecast(VAR_assault)
+
+autoplot(f_assault$forecast$assault) +
+  ggtitle("Prediction on how many more assault cases compared to yesterday") +
+  theme_minimal() +
+  ylab("Day-to-day change") +
+  xlab(paste("Number of days since", common_time[1]))
+
+```
+
+```{r r12, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vehicle thefts (left), and VAR model prediction for vehicle thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+# use car theft and criminal damage
+par(mfrow = c(1,2))
+lags = c(1:16)
+
+# criminal damange
+# only significant one is from covid to crime
+irf_criminal_2 <- irf(VAR_vehicle, 
+                         impulse = "residuals", 
+                         response = "vehicle", 
                          n.ahead = 15,
@@
   xlab("Days after a new COVID-19 case")+
-  ylab("Criminal Damage")+
+  ylab("Vehicle Theft")+
   theme(text = element_text(size=10), 
@@
 autoplot(f_criminal$forecast$criminal) +
-  ggtitle("Prediction on how many more criminal damage cases compared to yesterday") +
-  theme_minimal() +
-  ylab("Day-to-day change") +
-  xlab(paste("Number of days since", common_time[1]))
-
-```
-
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
-# use car theft and criminal damage
-par(mfrow = c(1,2))
-lags = c(1:16)
-
-# criminal damange
-# only significant one is from covid to crime
-irf_criminal_2 <- irf(VAR_vehicle, 
-                         impulse = "residuals", 
-                         response = "vehicle", 
-                         n.ahead = 15,
-                         ortho = F)
-
-# ggplot version of it.
-irf_criminal_2_gg <- data.frame(
-  irf_criminal_2$irf$residuals[,1],
-  irf_criminal_2$Lower$residuals[,1],
-  irf_criminal_2$Upper$residuals[,1]
-)
-
-colnames(irf_criminal_2_gg) <- c("mean", "lower", "upper")
-
-i1 <- ggplot(irf_criminal_2_gg, aes(x = lags)) +
-  geom_line(aes(y = mean), color = "black") +
-  geom_hline(yintercept = 0, color = "blue") +
-  theme_classic() +
-  ggtitle("") +
-  xlab("Days after a new COVID-19 case")+
-  ylab("Vehicle Theft")+
-  theme(text = element_text(size=10), 
-        plot.title = element_text(hjust = 0.5))
-i1
-
-interval_value_formatter <- "function(num, opts, seriesName, g, row, col) {
-  value = g.getValue(row, col);
-  if(value[0] != value[2]) {
-    lower = Dygraph.numberValueFormatter(value[0], opts);
-    upper = Dygraph.numberValueFormatter(value[2], opts);
-    return '[' + lower + ', ' + upper + ']';
-  } else {
-    return Dygraph.numberValueFormatter(num, opts);
-  }
-}"
-
-f_criminal <- forecast::forecast(VAR_criminal)
-
-autoplot(f_criminal$forecast$criminal) +
   ggtitle("Prediction on how many more Vehicle Theft cases compared to yesterday") +
@@
 ## Los Angeles
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
+
+When it comes to Los Angeles, only vandalism stood out based on the fitted VAR models.
+We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as battery ($p=0.67$), theft from vehicle ($p=0.9$), vehicle theft ($p=0.81$),  and burglary ($p=0.26$).
+Whereas, a statistically significant Granger causality was found between COVID-19 cases impacting vandalism ($p=0.003$).
+The impulse response of a new case to vandalism and the model predictions are outlined in Figure \@ref(fig:r21).
 
@@
 
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+```{r r21, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vandalism (left), and VAR model prediction for vandalism (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
 # use car theft and criminal damage
@@
 # criem significant to covid
-irf_battery_1 <- irf(VAR_battery,
-                     impulse = "battery",
-                     response = "residuals",
+irf_vandalism_1 <- irf(VAR_vandalism,
+                     impulse = "residuals",
+                     response = "vandalism",
                      n.ahead = 15)
 
-# html plot for battery
-irf_battery_1_gg <- data.frame(
-  irf_battery_1$irf$battery[,1],
-  irf_battery_1$Lower$battery[,1],
-  irf_battery_1$Upper$battery[,1]
+# html plot for vandalism
+irf_vandalism_1_gg <- data.frame(
+  irf_vandalism_1$irf$residuals[,1],
+  irf_vandalism_1$Lower$residuals[,1],
+  irf_vandalism_1$Upper$residuals[,1]
 )
 
-colnames(irf_battery_1_gg) <- c("mean", "lower", "upper")
-
-irf_battery_1_plot <- ggplot(irf_battery_1_gg, aes(x=lags)) +
+colnames(irf_vandalism_1_gg) <- c("mean", "lower", "upper")
+
+irf_vandalism_1_plot <- ggplot(irf_vandalism_1_gg, aes(x=lags)) +
   geom_line(aes(y = mean), color = "black") +
@@
   ggtitle("") +
-  xlab("Days after a new Battery case")+
-  ylab("COVID-19")+
+  xlab("Days after a new COVID-19 case")+
+  ylab("Vandalism")+
   theme(text = element_text(size=8), 
@@
 
-irf_battery_1_plot
-
-f_battery <- forecast::forecast(VAR_battery)
-
-autoplot(f_battery$forecast$battery) +
-  ggtitle("Prediction on how many more Battery cases compared to yesterday") +
+irf_vandalism_1_plot
+
+f_vandalism <- forecast::forecast(VAR_vandalism)
+
+autoplot(f_vandalism$forecast$vandalism) +
+  ggtitle("Prediction on how many more vandalism cases compared to yesterday") +
   theme_minimal() +
@@
 
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
-# use car theft and criminal damage
-lags = c(1:16)
-
-par(mfrow = c(1,2))
-
-# battery
-# criem significant to covid
-irf_burglary_1 <- irf(VAR_burglary,
-                     impulse = "burglary",
-                     response = "residuals",
-                     n.ahead = 15)
-
-# html plot for burglary
-irf_burglary_1_gg <- data.frame(
-  irf_burglary_1$irf$burglary[,1],
-  irf_burglary_1$Lower$burglary[,1],
-  irf_burglary_1$Upper$burglary[,1]
-)
-
-colnames(irf_burglary_1_gg) <- c("mean", "lower", "upper")
-
-irf_burglary_1_plot <- ggplot(irf_burglary_1_gg, aes(x=lags)) +
-  geom_line(aes(y = mean), color = "black") +
-  geom_hline(yintercept = 0, color = "blue") +
-  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "lightblue", alpha = 0.5) +
-  theme_classic() +
-  ggtitle("") +
-  xlab("Days after a new burglary case")+
-  ylab("COVID-19")+
-  theme(text = element_text(size=8), 
-        plot.title = element_text(hjust = 0.5))
-
-irf_burglary_1_plot
-
-f_burglary <- forecast::forecast(VAR_burglary)
-
-autoplot(f_burglary$forecast$burglary) +
-  ggtitle("Prediction on how many more burglary cases compared to yesterday") +
-  theme_minimal() +
-  ylab("Day-to-day change") +
-  xlab(paste("Number of days since", common_time[1]))
-
-```
-
-\newpage
-
 ## Philadelphia
 
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
+Coming to Philadelphia, only vehicle and general thefts stood out based on the fitted VAR models.
+We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as offense ($p=0.42$) and assault ($p=0.97$).
+On the contrary, a close statistically significant Granger causality was found between COVID-19 cases impacting vehicle theft ($p=0.0009$) and an instantaneous causality was found with thefts ($0.005$).
+The impulse response of a new case on these crimes and the model predictions are outlined in Figures \@ref(fig:r31) and \@ref(fig:r32).
 
@@
 
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+```{r r31, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on vehicle thefts (left), and VAR model prediction for vehicle thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
 
@@
 
-# general offense
+# general vehicle theft
 # irf
-irf_offense_1 <- irf(VAR_offense,
-                     impulse = "offense",
-                     response = "residuals",
+irf_vehicle_1 <- irf(VAR_vehicle,
+                     impulse = "residuals",
+                     response = "vehicle",
                      n.ahead = 15)
 # ggplot
-irf_offense_1_gg <- data.frame(
-  irf_offense_1$irf$offense[,1],
-  irf_offense_1$Lower$offense[,1],
-  irf_offense_1$Upper$offense[,1]
+irf_vehicle_1_gg <- data.frame(
+  irf_vehicle_1$irf$residuals[,1],
+  irf_vehicle_1$Lower$residuals[,1],
+  irf_vehicle_1$Upper$residuals[,1]
   )
 
-colnames(irf_offense_1_gg) <- c("mean", "lower", "upper")
-
-irf_offense_1_plot <- ggplot(irf_offense_1_gg, aes(x=lags)) +
+colnames(irf_vehicle_1_gg) <- c("mean", "lower", "upper")
+
+irf_vehicle_1_plot <- ggplot(irf_vehicle_1_gg, aes(x=lags)) +
   geom_line(aes(y = mean), color = "black") +
@@
   ggtitle("") +
-  xlab("Days after a New Other Offense Case")+
-  ylab("COVID-19")+
+  xlab("Days after a New COVID-19 Case")+
+  ylab("Vehicle Thefts")+
   theme(text = element_text(size=8), 
@@
 
-irf_offense_1_plot
-
-f_offense <- forecast::forecast(VAR_offense)
-
-autoplot(f_offense$forecast$offense) +
-  ggtitle("Prediction on how many more other offense cases compared to yesterday") +
+irf_vehicle_1_plot
+
+f_vehicle <- forecast::forecast(VAR_vehicle)
+
+autoplot(f_vehicle$forecast$vehicle) +
+  ggtitle("Prediction on how many more vehicle theft cases compared to yesterday") +
   theme_minimal() +
@@
 
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+```{r r32, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on thefts (left), and VAR model prediction for thefts (right).", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
 
@@
 irf_theft_1 <- irf(VAR_thefts,
-                   impulse = "thefts",
-                   response = "residuals",
+                   impulse = "residuals",
+                   response = "thefts",
                    n.ahead = 15)
@@
 irf_theft_1_gg <- data.frame(
-  irf_theft_1$irf$thefts[,1],
-  irf_theft_1$Lower$thefts[,1],
-  irf_theft_1$Upper$thefts[,1]
+  irf_theft_1$irf$residuals[,1],
+  irf_theft_1$Lower$residuals[,1],
+  irf_theft_1$Upper$residuals[,1]
 )
@@
   ggtitle("") +
-  xlab("Days after a new Theft case")+
-  ylab("COVID-19")+
+  xlab("Days after a new COVID-19 case")+
+  ylab("Thefts")+
   theme(text = element_text(size=8), 
@@
 ## Seattle
-Nulla consectetur erat arcu, eu porta urna sagittis nec.
-In sed vestibulum turpis.
-Nullam vel libero eu sapien eleifend condimentum.
-Donec nunc mi, ornare quis viverra ut, volutpat vitae orci.
-Quisque sit amet porttitor mauris, faucibus maximus arcu.
-Duis et lorem eu elit consectetur posuere.
-Aliquam blandit nisl nisl.
-Proin velit arcu, gravida vel tellus.
+
+Finally, for the city of Seattle, only burglaries stood out based on the fitted VAR models.
+We find that COVID-19 cases do not have a significant Granger instantaneous causality, at a 95% confidence level, with crimes such as larceny ($p=0.25$), vandalism ($p=0.79$), vehicle theft ($p=0.31$) and theft from vehicle ($p=0.46$).
+With a relaxed 90% confidence level, a statistically significant Granger causality was found between COVID-19 cases impacting burglaries ($p=0.073$).
+The impulse response of a new case on burglaries and the model predictions are outlined in Figure \@ref(fig:r41).
 
@@
 causality(VAR_larceny, cause = "residuals")
+
 causality(VAR_burglary, cause = colnames(combined_diff)[2])
@@
 
-```{r fig.show="hold",  fig.cap = "Frequency of each crime during months spanning the COVID-19 pandemic.
-Frequency of each crime during months spanning the COVID-19 pandemic.", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
+```{r r41, fig.show="hold",  fig.cap = "Impact of a new COVID-19 case on burglaries (left), and VAR model prediction for burglaries (right).
+", out.width="50%", message = FALSE, echo = FALSE, warning=FALSE}
 lags = c(1:16)
@@
   theme_classic() +
-  ggtitle("How many more burglary cases per day there will be 
-          after 1 confirmed covid19 case") +
   xlab("Number of days after a confimed COVID-19 case")+
@@
 ```
+
+Across all of the studied cities, we found that the crimes most significantly associated with COVID-19 cases are vehicle thefts and burglaries/thefts, with vandalism and assault also emerging as significant in some sectors.
+In the subsequent section we now discuss various implications of these findings and also discuss limitations of the study.
 

### 2022-05-01 outputs/paper/paper.Rmd
Included information about methods and also a detailed discussion on the results and limitations. Fixed Minor issues as well.

@@
 
-The key motivation for our work is previous work showing that certain crimes like domestic violence saw a significantly smaller drop as compared to other crimes after the onset of the pandemic [@bullinger2020covid], indicating the importance of the nature of the actual crime committed.
+The key motivation for our work is previous work showing that certain crimes like domestic violence saw a significantly smaller drop as compared to other crimes after the onset of the pandemic [[@boman2020has]], indicating the importance of the nature of the actual crime committed.
 Our focus on individual city data is due to known data collection issues for such research [@boman2020has] and is motivated by preliminary studies showing that the crime rates for specific crimes are largely dependent on the city [@ashby2020initial].
@@
 
-Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global, @campedelli2020disentangling, @nivette2021global] alongside work prompting further research in the area [@eisner2020violence].
+Overall, our paper is a small contribution building upon and motivated by several recent papers that have covered the impact of the pandemic on the human psyche and crime rates across different states and countries [@boman2021global; @campedelli2020disentangling; @nivette2021global] alongside work prompting further research in the area [@eisner2020violence].
 We achieve our goal of analyzing the pandemic data in conjunction with crime data obtained from the official city websites, using R statistical programming [@citeR], through time series modeling and also exploratory data analysis.
@@
 
+In this study we employ Vector autoregressive models to model the relationship between each crime and COVID-19 cases, which are then, in turn, used for Granger causality tests and finally impulse response modelling.
+
 ## Vector autoregression Model (VAR)
 
-Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over time.
-VAR is a type of stochastic process model.
-VAR models generalize the single-variable (univariate) autoregressive model by allowing for multivariate time series.
-VAR models are often used in economics and the natural sciences.
-A pth-order VAR is denoted "VAR(p)" and sometimes called "a VAR with p lags". A pth-order VAR model is written as
-
-$${\displaystyle y_{t}=c+A_{1}y_{t-1}+A_{2}y_{t-2}+\cdots +A_{p}y_{t-p}+e_{t},\,}y_{t}=c+A_{1}y_{{t-1}}+A_{2}y_{{t-2}}+\cdots +A_{p}y_{{t-p}}+e_{t},\,$$
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
+Vector autoregression (VAR) is a statistical model used to capture the relationship between multiple quantities as they change over a period of time.
+The main idea behind this model is to simply relate the present value for any variable at time $t$ to the past values of the variable.
+For our study, and each city, we fit a  VAR model between a given arbitrary crime denoted by $\text{CRIME}$ and the COVID-19 cases denoted by $\text{CASES}$.
+A pth-order VAR is fitted where p is decided based on a simpler generalised additive model.
+The equation then takes the following form,
+
+$$\text{CRIME}\displaystyle_{t}=c+A_{11}\text{CASES}_{t-1}+A_{12}\text{CASES}_{t-2}+\cdots +A_{1p}y_{t-p}+e_{1t},  $$
+
+$$\text{CASES}\displaystyle_{t}=c+A_{21}\text{CRIME}_{t-1}+A_{22}\text{CRIME}_{t-2}+\cdots +A_{2p}y_{t-p}+e_{2t}$$
+
+Here $A_i1 \ldots A_ip$ denote the coefficients for both the equations with $i\in \{1,2\}$ and $e_{it}$ denotes the error term.
+We fit these VAR models on the top five crimes for each city and then run Granger causality tests which are outlined below to establish causation.
+
+\bigskip
 
 ## Granger Causality Test
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.
-Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
-Proin commodo eros iaculis, faucibus dui et, vehicula felis.
-Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
-Mauris auctor sodales mauris, sed tristique quam sagittis in.
-Mauris ut bibendum enim.
-Maecenas consequat elit ut posuere aliquet.
-In hac habitasse platea dictumst.
-Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
-
+ To determine if COVID-19 cases in a specific locality are useful in forecasting a specific crime we employ Granger causality testing.
+The null-hypothesis for our specific test is simply that an arbitrary crime $\text{CRIME}$ is not caused by the the COVID-19 cases denoted by $\text{CASES}$.
+The test employs the formerly fitted VAR model to compute a $p$ value with the test-statistic here following a $F$ distribution.
+It should be noted here that this tests only for direct causality and not indirect causality which may occur due to a third unknown variable.
+
+Furthermore, we also conduct a simultaneous test for instantaneous Granger causality which is essentially a test that checks if both $x$ effects $y$ and $y$ effects $x$.
+The null hypothesis in this case is essentially that either  $x$ doesn't effect $y$ or $y$ doesn't effect $x$.
+More details about this test including the mathematical formulation are outlined in a much better fashion in a manuscript by @seth2007granger.
 <!-- \newpage -->
 
+\bigskip
+
 ## Forecasting and Impulse Response
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec egestas justo in nulla bibendum ultrices.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Donec porttitor interdum purus, vel commodo lorem lacinia at.
-Suspendisse posuere ex non risus dapibus, vitae ullamcorper ex sollicitudin.
-Vestibulum interdum augue sit amet tortor mollis, vitae ornare diam pulvinar.
-Proin sapien elit, tempor non consectetur ut, mollis quis nunc.
-Maecenas tincidunt velit dolor, eu porta lacus ornare et.
-Ut in cursus felis.
-Etiam porta, ipsum sit amet pulvinar consectetur, nulla metus vehicula quam, ac blandit lacus ex et ante.
-Aliquam lacinia sagittis lobortis.
-Integer eros felis, gravida nec augue ut, blandit ultricies ipsum.
-Morbi aliquet at magna rhoncus vehicula.
-Suspendisse aliquam risus nunc, vitae eleifend lacus venenatis ut.
-Sed nisi eros, malesuada nec purus ut, tristique congue dui.
-Maecenas velit massa, fringilla vel tristique vitae, condimentum ut purus.
-Proin commodo eros iaculis, faucibus dui et, vehicula felis.
-Integer ligula urna, feugiat quis tristique vel, cursus nec quam.
-Mauris auctor sodales mauris, sed tristique quam sagittis in.
-Mauris ut bibendum enim.
-Maecenas consequat elit ut posuere aliquet.
-In hac habitasse platea dictumst.
-Integer est arcu, semper ac dolor at, pulvinar vehicula mi.
+Finally, we also employ impulse response modelling that builds upon Granger tests to show us how a change of one standard deviation in $\text{CASES}$ does to the rate of  $\text{CRIME}$ in the future.
+Impulse response modelling is a simple process to investigate such relationships, specifically ones which are significant on the Granger test.
+We employ impulse response modelling to visualise these relationships and demonstrate how due to a single new COVID-19 case, specific crimes may fluctuate in the future.
+Further detail on Impulse response modelling has been outlined in @koop1996impulse.
+
+Next we investigate the criminal activity in each city using these modelling approaches and visualise the significant crimes in the subsequent section.
 
@@
 
-Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris ultricies felis in neque hendrerit semper.
-Sed quis ullamcorper nulla.
-Donec a mauris ut ipsum interdum tempor a et urna.
-Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
-Ut ullamcorper erat lorem, id congue ex ultricies ac.
-Nunc fermentum neque orci.
-Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
-Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
-Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
-Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
-Cras quam mauris, mollis a cursus non, cursus id neque.
-Cras eget elit posuere, ornare dui non, posuere neque.
-
-Ut tempor vestibulum augue sit amet hendrerit.
-Morbi rhoncus dignissim velit.
-Aliquam erat volutpat.
-Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
-Donec sit amet nibh in risus dictum venenatis.
-Mauris non volutpat risus, vitae ornare dui.
-Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
-Donec nec luctus quam.
-Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
-Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-Duis suscipit sem at enim commodo eleifend.
-Aenean eu malesuada justo.
-Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
-
-Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
-Nam commodo lectus odio, nec hendrerit purus consequat et.
-Sed vel iaculis ex.
-Maecenas mollis nisi id est pharetra facilisis.
-Fusce et tempor lorem.
-Etiam at facilisis nulla.
-Praesent sed erat odio.
-Integer sed dictum tellus, in finibus odio.
-Duis et rhoncus justo.
-Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris ultricies felis in neque hendrerit semper.
-Sed quis ullamcorper nulla.
-Donec a mauris ut ipsum interdum tempor a et urna.
-Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
-Ut ullamcorper erat lorem, id congue ex ultricies ac.
-Nunc fermentum neque orci.
-Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
-Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
-Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
-Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
-Cras quam mauris, mollis a cursus non, cursus id neque.
-Cras eget elit posuere, ornare dui non, posuere neque.
-
-Ut tempor vestibulum augue sit amet hendrerit.
-Morbi rhoncus dignissim velit.
-Aliquam erat volutpat.
-Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
-Donec sit amet nibh in risus dictum venenatis.
-Mauris non volutpat risus, vitae ornare dui.
-Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
-Donec nec luctus quam.
-Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
-Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-Duis suscipit sem at enim commodo eleifend.
-Aenean eu malesuada justo.
-Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
-
-Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
-Nam commodo lectus odio, nec hendrerit purus consequat et.
-Sed vel iaculis ex.
-Maecenas mollis nisi id est pharetra facilisis.
-Fusce et tempor lorem.
-Etiam at facilisis nulla.
-Praesent sed erat odio.
-Integer sed dictum tellus, in finibus odio.
-Duis et rhoncus justo.
-Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Mauris ultricies felis in neque hendrerit semper.
-Sed quis ullamcorper nulla.
-Donec a mauris ut ipsum interdum tempor a et urna.
-Ut vehicula ligula sit amet nunc sodales, et facilisis felis vestibulum.
-Ut ullamcorper erat lorem, id congue ex ultricies ac.
-Nunc fermentum neque orci.
-Quisque congue, ligula eget tincidunt condimentum, nulla leo suscipit nibh, eu auctor tellus augue nec turpis.
-Nulla lorem dui, scelerisque non molestie et, pharetra in enim.
-Curabitur placerat leo a purus pellentesque, lobortis hendrerit justo posuere.
-Fusce condimentum porttitor purus, eget ullamcorper elit bibendum et.
-Cras quam mauris, mollis a cursus non, cursus id neque.
-Cras eget elit posuere, ornare dui non, posuere neque.
-
-Ut tempor vestibulum augue sit amet hendrerit.
-Morbi rhoncus dignissim velit.
-Aliquam erat volutpat.
-Maecenas semper nulla nec sem tempus, eget elementum odio pulvinar.
-Donec sit amet nibh in risus dictum venenatis.
-Mauris non volutpat risus, vitae ornare dui.
-Praesent blandit leo nec dolor convallis, ut convallis lacus bibendum.
-Donec nec luctus quam.
-Nam congue suscipit ex, sed suscipit ligula vestibulum vel.
-Orci varius natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-Duis suscipit sem at enim commodo eleifend.
-Aenean eu malesuada justo.
-Sed finibus tincidunt lectus, eget facilisis leo porttitor id.
-
-Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
-Nam commodo lectus odio, nec hendrerit purus consequat et.
-Sed vel iaculis ex.
-Maecenas mollis nisi id est pharetra facilisis.
-Fusce et tempor lorem.
-Etiam at facilisis nulla.
-Praesent sed erat odio.
-Integer sed dictum tellus, in finibus odio.
-Duis et rhoncus justo.Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
-
-Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
-Nam commodo lectus odio, nec hendrerit purus consequat et.
-Sed vel iaculis ex.
-Maecenas mollis nisi id est pharetra facilisis.
-Fusce et tempor lorem.
-Etiam at facilisis nulla.
-Praesent sed erat odio.
-Integer sed dictum tellus, in finibus odio.
-Duis et rhoncus justo.Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
-
-Morbi pellentesque ligula sed orci lacinia, in vulputate velit vestibulum.
-Nam commodo lectus odio, nec hendrerit purus consequat et.
-Sed vel iaculis ex.
-Maecenas mollis nisi id est pharetra facilisis.
-Fusce et tempor lorem.
-Etiam at facilisis nulla.
-Praesent sed erat odio.
-Integer sed dictum tellus, in finibus odio.
-Duis et rhoncus justo.Pellentesque id purus eros.
-Curabitur rutrum ut nisl vel facilisis.
-Maecenas dapibus vel justo ut aliquet.
-Suspendisse at auctor sem.
-Pellentesque sit amet tortor vel arcu sollicitudin lacinia.
-Pellentesque condimentum, metus eget maximus aliquam, neque nisl.
+## Overview of Results
+
+During the COVID-19 pandemic, the world has seen several different changes both negative and positive.
+With increasing stay-at-home restrictions and various other policy measures, we aim to examine the impact that the pandemic has on criminal activity across several different cities in the United States including Chicago, Los Angeles, Seattle, and Philadelphia.
+The main goal of this study is to take a closer look at how different crimes varied across the pandemic as well as compare them to trends prior to the pandemic.
+
+Consistent with early research that anticipated an increase in thefts and burglaries with the onset of the pandemic [@miller2020crime], our study found that there was a strong correlation between the rate of COVID-19 cases and crimes such as thefts and burglaries, especially in cities such as Seattle and Philadelphia, as seen in Figures \@ref(fig:r41) and \@ref(fig:r32).
+There was also a significant relationship between vehicle thefts and COVID-19 cases in cities such as Philadelphia and Chicago.
+as seen in Figures \@ref(fig:r31) and \@ref(fig:r12).
+Overall, our exploratory analysis also indicated similar trends, with the uptick in thefts and burglaries possibly occurring due to the lack of guardianship of assets due to people staying home.
+Furthermore, our study indicated that crimes such as vandalism also show some relation with COVID-19 cases when it comes to cities like L.A, as seen in Figure \@ref(fig:r21), alongside other crimes such as assault in cities like Chicago, as observed in Figure \@ref(fig:r11).
+Overall it is noted that crimes that are better committed individually such as assault, theft, and burglary stood out more as compared to crimes committed in groups like vandalism.
+This is consistent with other research on crime in the United States which arrived at similar conclusions [@boman2020has].
+
+The impulse response graphs show us another perspective with respect to these crimes.
+These graphs show us the future effects of a new COVID-19 case on a specific criminal activity.
+Based on the Figures  \@ref(fig:r12),  \@ref(fig:r31),  \@ref(fig:r32) &  \@ref(fig:r41) it can be noted that on average across all of the cities examined there is a high likelihood of at least one vehicle theft or burglary occurring after a unit increase in cases.
+Furthermore, based on our exploratory analysis carried out in Section 2, there seems to be no significant change in the frequency of occurrence of several crimes across several cities, pointing to a more bleak reality as compared to prior studies indicating an overall crime reduction.
+
+These results in a way bring into question the significance of the overall decline in crime as there seem to be crimes like thefts and vehicle thefts that seem to have gone up significantly when compared to historical data.
+Furthermore, it seems like if the pandemic worsens in the future, we might see more crimes that are simply caused due to lack of guardianship as pointed out earlier.
+However, before arriving at such concrete conclusions we must take into consideration some of the limitations of this study and also explore future directions.
+
+## Ethical Concerns and Limitations
+
+Across studies that have examined crime during the COVID-19 pandemic, the lack of sufficient data has been a common limitation [@boman2020has; @ashby2020initial; @boman2021global; @campedelli2020disentangling; @nivette2021global].
+In our study too, we find that unification of several datasets derived from different police units and cities is hard, and arriving at a concrete common conclusion based on them is even harder.
+This study looked at data collected, cleaned, and stored by different organizations across the United States impacted in different ways by both crime and criminal activity, making it important to acknowledge that the conclusions we arrive at might not be truly generalizable.
+
+When it comes to our modeling approach, we note that the VAR models used are quite limited in their ability as the only variables involved in our models are counts of the specific crime and COVID-19 cases.
+These models should not be used to forecast, COVID-19 or Crime cases as there are several other factors at play that influence either of their occurrences.
+Another major limitation and concern are to note that the Granger causality does not determine true causality and is only one of the many metrics that can be used to determine the cause-effect relationship between variables.
+The Granger test fails when both the involved variables are simultaneously affected by a third unknown variable, thus giving the illusion of causality.
+We only use the Granger test in our study to narrow down relevant variables but future work can certainly employ other more sophisticated Bayesian Causality metrics like GARCH  [@chen2017bayesian].
+Finally, there are also several biases like underreporting of certain types of crimes and racial bias that plague crime data [@hetey2018numbers; @buil2021accuracy] that have to be acknowledged as well.
+It should also be noted that, due to stay-at-home measures and lack of access to police stations or support, several crimes especially rape and sexual assault are even more underreported than usual limiting the scope of our research [@palermo2011undercounting].
+
+## Future Directions
+
+Finally, our study opens the door for more future work exploring the impact brought upon by the COVID-19 pandemic on crime and the human psyche.
+With  the now availability of more comprehensive datasets like the global crime data set compiled by @nivette2021global, such an analysis can be further extended to other countries.
+With more sophisticated modelling approaches such as machine learning inspired recurrent neural networks to make more accurate predictions, our predictive approaches can be made more expressive and further improved to arrive at better informed conclusions.
+A more comprehensive study can also include other crimes such as domestic violence and hate crimes against racial minorities alongside cybercrime which seems to have increased over the pandemic due to increased internet use.
 
